### 2015-02-08 rmd/paper.Rmd
add mortarboard/rmd version

---
title: 'Claret: Using Data Types for Highly Concurrent Distributed Transactions'
preprint: true

author:
  - {family: Holt,  given: Brandon, affiliation: 1, email: bholt}
  - {family: Zhang, given: Irene,   affiliation: 1, email: iyzhang}
  - {family: Ports, given: Dan,     affiliation: 1, email: dkp}
  - {family: Oskin, given: Mark,    affiliation: 1, email: oskin}
  - {family: Ceze,  given: Luis,    affiliation: 1, email: luisceze}

organization:
  - {id: 1, name: University of Washington}

conference:
  name: PaPoC 2015
  location: 'April 21, 2015, Bordeaux, France'
  year: 2015

doi: 0

layout: sigplanconf
bibliography: biblio.bib
output:
  pdf_document:
    fig_caption: yes

abstract: |
  Building database applications out of data structures rather than simple string values allows the flexibility and fine-grained control of typical key-value databases while providing better performance and scalability. Composing transactions out of linearizable data structure operations exposes concurrency in a safe way, making it simple to implement efficiently and easy to reason about.
---
```{r setup, include=FALSE}
opts_chunk$set(dev = 'pdf')
```

# Introduction

Providing strong consistency in interactions with databases greatly aids programmers in their ability to write correct code. However, when scaling services under heavy load to millions or billions of users, systems today often must give up these consistency guarantees to meet tight performance goals. A wide variety of NoSQL databases support a more relaxed, or eventually consistent model, and in the majority of cases, it seems that this lack of consistency does not hinder them — the observation being that most times, transactional updates are unneccessary because potential conflicts are just not likely to be observed, and the users may even accept minor inconsistencies, such as two tweets being out of temporal order.

However, this is leaving a lot to chance. If it truly is the case that these interactions shouldn't conflict in observable ways, or that certain parts of the application can tolerate imprecision, then why not capture those properties in the programming model for these databases? What if the semantics of operations were known to the database, so it could continue to ignore cases where ordering or consistency really don't matter, but could step in and mediate those specific instances where it could be a problem. And what if the programmer could express the semantics they desire succintly and rigorously?

We argue that the correct way to express and enforce these semantics is through the use of an old computer science standby: abstract data types (ADTs). Rather than treating the values in a key/value store simply as strings, making them into more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a clean mechanism for programmers to express their intentions.

Performance benefits come from understanding the properties of ADT operations: those that commute with each other can be performed concurrently, even on multiple copies of the record. This means that transactions whose operations commute abort less, approaching the performance without transactions. This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control. Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.

Selecting the right data type for the job gives programmers a clean, precise way to express their desired behavior. For instance, rather than using a generic integer to generate unique identifiers, a `UniqueID` type, realizing that contiguity isn't necessary, can be trivially parallelized and distributed. Though this is an extremely simple case (and nearly universally adopted optimization), it fits the same mold as more nuanced decisions, such as choosing to represent the number of times a given post has been "retweeted" as a `HyperLogLog`, which can efficiently yield the approximate number of unique users, rather than a costly precise `Set`. Though selecting data structures for the job at hand is nothing new to programmers, only a handful of databases, such as Redis, MongoDB or Riak, support this flexibility, and they do not use the abstraction it affords to enforce strongly consistent transactions.

# Commutativity {#comm}

Commutativity is well known, especially in distributed systems, for enabling important optimizations. Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict. Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.

In the realm of eventual consistency, commutativity has been leveraged for convergence guarantees.
RedBlue consistency allows executing commutative ("blue") operations locally, knowing they will eventually converge. Similarly, conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11} define commutative merge functions for all operations to ensure that replicas will converge.

Lynx \cite{Zhang:SOSP13} uses knowledge of some commutative operations to make tracking serializability cheaper in chained transactions. Doppel \cite{Narula:OSDI14} added several explicitly commutative operations on records which they exploited to better handle common high-contention situations such as counters and "top-k lists" in the context of a single node multicore database. Finally, HyFlow \cite{Kim:EuroPar13}, a distributed transactional memory framework, reorders commutative operations on specific data types to execute before others to allow them to operate concurrently on a single version of a record.

## Commutativity Specifications

\begin{table}
\centering
\begin{tabular}{lll}
\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
\hline
\texttt{add(x) -> void} & \texttt{add(y)} & $\forall x, y$ \\
\texttt{remove(x) -> void} & \texttt{remove(y)} & $\forall x, y$ \\
    & \texttt{add(y)} & $x \ne y$ \\
\texttt{size() -> int} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
\texttt{getall() -> list} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
    & \texttt{size()} & $always$ \\
    \hline
\end{tabular}
\caption{\label{spec} Commutativity Specification for Set.}
\end{table}

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{tab:spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{tab:spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

## Transactional Boosting
If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record. This technique of raising the level of abstraction in transactions to operations is known as *transactional boosting* \cite{Herlihy:PPoPP08}. This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory. In \autoref{eval}, we show how it can be applied to distributed transactions.

## Other opportunities

Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations, even without aborting, is a bottleneck which prevents scaling. The Ellen Degeneres selfie retweet is a prime example of this. In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity. We mention just a few here which we have yet to evaluate.

### Record splitting for parallelism

Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation. This allows them to *split* hot records onto multiple cores and execute commutative operations in parallel, *reconciling* them back to a single value before executing reads or non-commuting operations. This observation can be generalized via ADTs: operations can operate on copies of a record in parallel provided they commute with each other and provided they can be combined at the end of a phase before beginning to execute a new phase with a different set of operations that didn't commute with the first set.

### Combining

Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself. This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.

# Data type selection

Choosing the right data type for your application can have profound performance implications.
As mentioned before, the interface can determine what commutes simply by exposing more information than necessary. By selecting the semantics that are most permissive or specialized for their particular use case, programmers can ensure that the system is given the best chance of scaling performance effectively. By allowing approximations or non-determinism, performance may be further improved.

## Unique identifiers
As an extremely simple motivating example, imagine a programmer wants a unique identifier for each new post. They might naively choose a `Counter` which returns the next number in the sequence whenever `next` is called. Ensuring each caller gets the next value without skipping any ends up being very expensive, as implementers of TPC-C \cite{TPCC}, which explicitly requires this, know well. In most cases, however, the programmer could use a special `UniqueID` type, whose only criteria is that each call to `next` returns a different value, which can be implemented trivially in a number of efficient highly-concurrent ways.

## Probabilistic data types

Some data types have *probabilistic* guarantees about their semantics, which, rather than always returning a precisely correct answer, trade off some accuracy for better performance or storage. Some better-known examples include *bloom filters*\cite{bloom}, *hyperloglog* \cite{hyperloglog}, and *count-min sketch* \cite{countminsketch}. Hyperloglog, which also appears in Redis \cite{redis}, estimates the cardinality (size) of a set within a fixed error bound. Twitter's streaming analytics system \cite{summingbird} leverages these probabilistic data types to handle the high volume of data needing to be processed. We expect similar improvements to be had from their use in Claret.

## Conflict-free replicated data types

CRDTs, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type with loosly synchronized semantics. These data types might allow multiple copies in different shards to asynchronously update each other on the operations they've seen. By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state. For instance, a Set CRDT must decide what to do when two clients concurrently add and remove the same key — usually this is handled by choosing a preference for adds. Clients of this data type might find it more difficult to reason about, since a remove may appear to never have happened, but one can imagine making this tradeoff in parts of the application where it otherwise cannot scale.

This mixing in of non-serializability should not be taken lightly. To ensure that serializable transactions aren't tainted by inconsistent information from these kinds of data types, some form of information flow or static analysis (via a type system) could be employed in client code.

# Evaluation {#eval}
To show the efficacy of leveraging commutative operations, we use an application typical of web workloads: a simplified Twitter clone known as *Retwis*. In \autoref{tput} you can see that with commutativity, transaction throughput scales with increased concurrency.

```{r throughput, fig.cap="Throughput on uniform random workload.\\label{tput}", fig.width=3.5, fig.height=3, echo=F, message=F, warning=F, error=F}
data <- function(d) {
  d$abort_rate <- d$txn_failed / (d$txn_count + d$txn_failed)
  d$throughput <- d$txn_count * num(d$nclients) / d$total_time
  # d$throughput <- d$ntxns * num(d$nclients) / d$total_time
  d$avg_latency_ms <- d$txn_time / d$txn_count * 1000
  return(d)
}

d <- data(db("
  select * from tapir where 
  generator_time is not null and total_time is not null
  and (initusers = 50 or initusers = 500)
  and name like 'claret-v%'
",
  factors=c('nshards', 'nclients'),
  numeric=c('total_time', 'txn_count')
))

ggplot(d, aes(
  x = nclients,
  y = throughput,
  group = ccmode,
  fill = ccmode,
  color = ccmode
))+
# geom_meanbar()+
stat_smooth()+
facet_grid(~nshards, labeller=label_pretty)+
theme_mine
```

### 2015-02-09 rmd/paper.Rmd
rmd: update styling & spec

@@
 \hline
-\texttt{add(x) -> void} & \texttt{add(y)} & $\forall x, y$ \\
-\texttt{remove(x) -> void} & \texttt{remove(y)} & $\forall x, y$ \\
+\texttt{add(x): void} & \texttt{add(y)} & $\forall x, y$ \\
+\texttt{remove(x): void} & \texttt{remove(y)} & $\forall x, y$ \\
     & \texttt{add(y)} & $x \ne y$ \\
-\texttt{size() -> int} & \texttt{add(x)} & $x \in Set$ \\
+\texttt{size(): int} & \texttt{add(x)} & $x \in Set$ \\
     & \texttt{remove(x)} & $x \notin Set$ \\
-\texttt{getall() -> list} & \texttt{add(x)} & $x \in Set$ \\
-    & \texttt{remove(x)} & $x \notin Set$ \\
-    & \texttt{size()} & $always$ \\
+\texttt{contains(x): bool} & \texttt{add(y)} & $x \ne y \lor y \in Set$ \\
+    & \texttt{remove(y)} & $x \ne y \lor y \notin Set$ \\
+    & \texttt{size()} & $\forall x$ \\
     \hline
@@
 Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
-We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{tab:spec}.
+We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
 It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$.
@@
 
-For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{tab:spec}, but there may be many that express varying degrees of commutativity.
+For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity.
 The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute.
@@
 
-### Record splitting for parallelism
-
+**Record splitting for parallelism.**
 Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation.
@@
 
-### Combining
-
+**Combining.**
 Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself.

### 2015-02-14 rmd/paper.Rmd
incorporate latest plots: update throughput & add stress throughput

@@
 ```{r setup, include=FALSE}
-opts_chunk$set(dev = 'pdf')
+opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.2, fig.height=3)
 ```
@@
 
-```{r throughput, fig.cap="Throughput on uniform random workload.\\label{tput}", fig.width=3.5, fig.height=3, echo=F, message=F, warning=F, error=F}
-data <- function(d) {
-  d$abort_rate <- d$txn_failed / (d$txn_count + d$txn_failed)
-  d$throughput <- d$txn_count * num(d$nclients) / d$total_time
-  # d$throughput <- d$ntxns * num(d$nclients) / d$total_time
-  d$avg_latency_ms <- d$txn_time / d$txn_count * 1000
-  return(d)
-}
+```{r stress, fig.cap="Throughput of SortedSet microbenchmark.\\label{stress_tput}"}
+d <- data(db("
+    select * from tapir where 
+    total_time is not null
+    and name like 'stress-v0.14%'
+    ", factors=c('nshards', 'nclients'), numeric=c('total_time', 'txn_count')))
+d$opmix <- factor(revalue(d$mix, c(
+    'mostly_update'='35% read\n65% update',
+    'update_heavy'='50% read\n50% update',
+    'read_heavy'='90% read\n10% update'
+)))
+d.u <- subset(d, nshards == 4 & nkeys == 10000 & alpha == 0.6 & grepl('update_heavy|read_heavy', mix))
+ggplot(d.u, aes(x=nclients, y=throughput, group=cc, fill=cc, color=cc, linetype=cc))+
+    stat_summary(fun.y=mean, geom="line")+
+    xlab('Concurrent clients')+ylab('Throughput (transactions / sec)')+
+    expand_limits(y=0)+
+    facet_wrap(~opmix)+
+    theme_mine+theme(legend.position='top', legend.direction='horizontal', legend.title.align=1)+
+    cc_scales(title='Concurrency\ncontrol:')
+```
 
+```{r throughput, fig.cap="Throughput on social network workloads.\\label{tput}"}
 d <- data(db("
@@
     generator_time is not null and total_time is not null
-  and (initusers = 50 or initusers = 500)
-  and name like 'claret-v%'
+    and name like 'claret-v0.14%'
   ",
@@
 ))
-
-ggplot(d, aes(
-  x = nclients,
-  y = throughput,
-  group = ccmode,
-  fill = ccmode,
-  color = ccmode
-))+
-# geom_meanbar()+
-stat_smooth()+
-facet_grid(~nshards, labeller=label_pretty)+
-theme_mine
+d.u <- subset(d, nshards == 4 & initusers == 4096 & grepl('geom_repost|read_heavy', mix))
+ggplot(d.u, aes(x=nclients, y=throughput, group=cc, fill=cc, color=cc, linetype=cc))+
+    stat_summary(fun.y=mean, geom="line")+
+    xlab('Concurrent clients')+ylab('Throughput (transactions / sec)')+
+    expand_limits(y=0)+
+    facet_wrap(~workload)+
+    theme_mine+theme(legend.position='top', legend.direction='horizontal', legend.title.align=1)+
+    cc_scales(title='Concurrency\ncontrol:')
 ```

### 2015-02-14 rmd/paper.Rmd
port over the latest from Authorea to rmd

@@
 abstract: |
-  Building database applications out of data structures rather than simple string values allows the flexibility and fine-grained control of typical key-value databases while providing better performance and scalability.
-Composing transactions out of linearizable data structure operations exposes concurrency in a safe way, making it simple to implement efficiently and easy to reason about.
+    Out of the many NoSQL databases in use today, some that provide simple data structures for records, such as Redis and MongoDB, are now becoming popular.
+Building applications using these data structures rather than plain string values provides programmers with a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema.
+Currently, this expressiveness is used to ensure related values are co-located so they can be quickly accessed together (e.g.
+maps and other aggregates) and to provide complex atomic operations (e.g.
+set insertion).
+However, there are many more ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.
+
+    In this work, we demonstrate several ways of leveraging data structure semantics in databases, focusing primarily on commutativity.
+Reasoning about operation reordering can allow transactions to execute concurrently that would conflict under traditional concurrency control.
+Using Retwis, a Twitter clone built for Redis, as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks.
+We conclude that data types are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
 ---
@@
 
-Providing strong consistency in interactions with databases greatly aids programmers in their ability to write correct code.
-However, when scaling services under heavy load to millions or billions of users, systems today often must give up these consistency guarantees to meet tight performance goals.
-A wide variety of NoSQL databases support a more relaxed, or eventually consistent model, and in the majority of cases, it seems that this lack of consistency does not hinder them — the observation being that most times, transactional updates are unneccessary because potential conflicts are just not likely to be observed, and the users may even accept minor inconsistencies, such as two tweets being out of temporal order.
+The move to non-relational (NoSQL) databases was motivated by a desire for scalability and flexibility.
+People found that by not enforcing strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
+By forgoing the structure of relational schemas, they gained more direct control over performance at the cost of significant automatic optimization.
+The observation has been that relaxed or eventual consistency is a good fit for many applications built on top of these NoSQL because potential conflicts are difficult to observe in practice and that users are likely to accept minor inconsistencies, such as two tweets being out of temporal order on their Twitter timeline.
+However, this leaves much to chance; there is likely no guarantee that more significant inconsistencies are impossible.
 
-However, this is leaving a lot to chance.
-If it truly is the case that these interactions shouldn't conflict in observable ways, or that certain parts of the application can tolerate imprecision, then why not capture those properties in the programming model for these databases?
-What if the semantics of operations were known to the database, so it could continue to ignore cases where ordering or consistency really don't matter, but could step in and mediate those specific instances where it could be a problem.
-And what if the programmer could express the semantics they desire succintly and rigorously?
+In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong.
+Google, recognizing the benefit to programmer productivity, reintroduced serializable transactions to their next generation database, Spanner \cite{Spanner}.
+However, this leaves them with two extreme choices, with significant performance impact.
+If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model?
+What if the programmer could express the semantics they desire succintly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?
 
-We argue that the correct way to express and enforce these semantics is through the use of an old computer science standby: abstract data types (ADTs).
-Rather than treating the values in a key/value store simply as strings, making them into more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a clean mechanism for programmers to express their intentions.
+Abstract data types (ADTs) are the solution to this problem.
+Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
 
@@
 This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control.
-Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.
+Using insights from multi-core concurrent data structures, we show in Section \ref{sec:commutativity} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.
 
 Selecting the right data type for the job gives programmers a clean, precise way to express their desired behavior.
-For instance, rather than using a generic integer to generate unique identifiers, a `UniqueID` type, realizing that contiguity isn't necessary, can be trivially parallelized and distributed.
+For instance, rather than using a generic integer to generate unique identifiers, a `UUID` type, realizing that contiguity of ids isn't necessary, can be trivially parallelized and distributed.
 Though this is an extremely simple case (and nearly universally adopted optimization), it fits the same mold as more nuanced decisions, such as choosing to represent the number of times a given post has been "retweeted" as a `HyperLogLog`, which can efficiently yield the approximate number of unique users, rather than a costly precise `Set`.
@@
 # Commutativity {#comm}
+\label{sec:commutativity}
 
@@
 Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
-We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
+We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in Table \ref{tab:spec}.
 It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$.
@@
 
-For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity.
+For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in Table \ref{tab:spec}, but there may be many that express varying degrees of commutativity.
 The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute.
@@
 This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory.
-In \autoref{eval}, we show how it can be applied to distributed transactions.
+In Section \ref{sec:evaluation}, we show how it can be applied to distributed transactions.
 
@@
 
-**Record splitting for parallelism.**
+### Record splitting for parallelism
+
 Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation.
@@
 
-**Combining.**
+### Combining
+
 Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself.
 This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.
-
-# Data type selection
-
-Choosing the right data type for your application can have profound performance implications.
-As mentioned before, the interface can determine what commutes simply by exposing more information than necessary.
-By selecting the semantics that are most permissive or specialized for their particular use case, programmers can ensure that the system is given the best chance of scaling performance effectively.
-By allowing approximations or non-determinism, performance may be further improved.
-
-## Unique identifiers
-As an extremely simple motivating example, imagine a programmer wants a unique identifier for each new post.
-They might naively choose a `Counter` which returns the next number in the sequence whenever `next` is called.
-Ensuring each caller gets the next value without skipping any ends up being very expensive, as implementers of TPC-C \cite{TPCC}, which explicitly requires this, know well.
-In most cases, however, the programmer could use a special `UniqueID` type, whose only criteria is that each call to `next` returns a different value, which can be implemented trivially in a number of efficient highly-concurrent ways.
-
-## Probabilistic data types
-
-Some data types have *probabilistic* guarantees about their semantics, which, rather than always returning a precisely correct answer, trade off some accuracy for better performance or storage.
-Some better-known examples include *bloom filters*\cite{bloom}, *hyperloglog* \cite{hyperloglog}, and *count-min sketch* \cite{countminsketch}.
-Hyperloglog, which also appears in Redis \cite{redis}, estimates the cardinality (size) of a set within a fixed error bound.
-Twitter's streaming analytics system \cite{summingbird} leverages these probabilistic data types to handle the high volume of data needing to be processed.
-We expect similar improvements to be had from their use in Claret.
-
-## Conflict-free replicated data types
-
-CRDTs, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type with loosly synchronized semantics.
-These data types might allow multiple copies in different shards to asynchronously update each other on the operations they've seen.
-By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state.
-For instance, a Set CRDT must decide what to do when two clients concurrently add and remove the same key — usually this is handled by choosing a preference for adds.
-Clients of this data type might find it more difficult to reason about, since a remove may appear to never have happened, but one can imagine making this tradeoff in parts of the application where it otherwise cannot scale.
-
-This mixing in of non-serializability should not be taken lightly.
-To ensure that serializable transactions aren't tainted by inconsistent information from these kinds of data types, some form of information flow or static analysis (via a type system) could be employed in client code.
 

### 2015-02-14 rmd/paper.Rmd
write evaluation section

@@
 # Evaluation {#eval}
-To show the efficacy of leveraging commutative operations, we use an application typical of web workloads: a simplified Twitter clone known as *Retwis*.
-In \autoref{tput} you can see that with commutativity, transaction throughput scales with increased concurrency.
+To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of available operations.
+The transaction protocol employs fairly standard two-phase commit design, and uses two-phase locking with retries to guarantee isolation.
+To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
+During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
+an "increment" speculatively returns the incremented value).
+When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
+This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
 
-```{r stress, fig.cap="Throughput of SortedSet microbenchmark.\\label{stress_tput}"}
+Commutativity comes into play in the locking scheme.
+Using the algorithms from \cite{Kulkarni:PLDI11} and our commutativity specifications, we design an abstract lock for each record type.
+Our `SortedSet`, for instance, has an `add` mode which allows all insertion operations to commute, but disallows operations like `contains` or `size`.
+As a baseline, we implement a standard reader/writer locking scheme that allows all read-only operations to execute concurrently, but enforces that only one transaction may modify a record at a time.
+
+These experiments were run with 4 shards on 4 local nodes, each with 8-core 2GHz Xeon E5335 processors.
+
+```{r stress, fig.cap="Throughput of raw Set operations., \\label{stress_tput}", fig.height=3.3}
 d <- data(db("
@@
 )))
-d.u <- subset(d, nshards == 4 & nkeys == 10000 & alpha == 0.6 & grepl('update_heavy|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput, group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=mean, geom="line")+
-    xlab('Concurrent clients')+ylab('Throughput (transactions / sec)')+
+d$dist <- factor(revalue(d$alpha, c('0.6'='Zipf: 0.6', '-1'='Uniform')))
+d.u <- subset(d, nshards == 4 & nkeys == 10000 & (alpha == '0.6' | alpha == '-1') & grepl('update_heavy|read_heavy', mix))
+ggplot(d.u, aes(x=nclients, y=throughput/1000, group=cc, fill=cc, color=cc, linetype=cc))+
+    stat_summary(fun.y=max, geom="line")+
+    xlab('Concurrent clients')+ylab('Throughput (k/sec)')+
     expand_limits(y=0)+
-    facet_wrap(~opmix)+
-    theme_mine+theme(legend.position='top', legend.direction='horizontal', legend.title.align=1)+
+    facet_grid(dist~opmix)+
+    theme_mine+theme(legend.position='bottom', legend.direction='horizontal', legend.title.align=1)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 
-```{r throughput, fig.cap="Throughput on social network workloads.\\label{tput}"}
+**Microbenchmark: Set operations.** We first evaluate performance with a simple workload consisting of a raw mix of `Set` operations randomly distributed over 10,000 keys.
+We use both a uniform random distribution as well as a skewed Zipfian distribution with a coefficient of 0.6.
+In \autoref{stress_tput}, we see that commutative transactions perform strictly better, showing the most pronounced benefit over the more update-heavy, skewed workload.
+
+```{r throughput, fig.cap="Throughput on social network workload (Retwis).
+Leveraging commutativity prevents performance from falling over even when a posts spread virally across the network (repost-heavy).\\label{tput}"}
 d <- data(db("
@@
     facet_wrap(~workload)+
-    theme_mine+theme(legend.position='top', legend.direction='horizontal', legend.title.align=1)+
+    theme_mine+theme(legend.position='bottom', legend.direction='horizontal', legend.title.align=1)+
     cc_scales(title='Concurrency\ncontrol:')
 ```
+
+**Case study: Retwis.** To understand performance of more typical of web workloads, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis.
+We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
+On top of Retwis's basic functionality, we also add a "repost" option that works like Twitter's "retweet".
+We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
+More about this workload is in \autoref{apx:retwis}.
+
+\autoref{tput} shows the results of this simulation.
+When most of the traffic is content consumption (reading timelines), both systems perform well enough.
+However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
+As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over.
+But with commutativity, we see that performance continues to scale even under this arduous load.

### 2015-02-14 rmd/paper.Rmd
appendix

@@
 # Evaluation {#eval}
-To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of available operations.
-The transaction protocol employs fairly standard two-phase commit design, and uses two-phase locking with retries to guarantee isolation.
-To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
-During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
-an "increment" speculatively returns the incremented value).
-When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
-This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
+To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations.
+The transaction protocol employs standard two-phase commit, and two-phase locking, more details can be found in the appendix (\autoref{apx:protocol}).
 
@@
 
-```{r throughput, fig.cap="Throughput on social network workload (Retwis).
-Leveraging commutativity prevents performance from falling over even when a posts spread virally across the network (repost-heavy).\\label{tput}"}
+```{r throughput, fig.cap="Throughput of social network workload (Retwis) with 4000 users.
+Leveraging commutativity prevents performance from falling over even when posts spread virally (repost-heavy).\\label{tput}"}
 d <- data(db("
@@
 
-**Case study: Retwis.** To understand performance of more typical of web workloads, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis.
+**Case study: Retwis.** 
+To understand performance of more typical of web workloads, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis.
 We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
@@
 We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
-More about this workload is in \autoref{apx:retwis}.
+More about this workload is in the appendix (\autoref{apx:retwis}).
 
@@
 But with commutativity, we see that performance continues to scale even under this arduous load.
+
+# Appendix
+
+## Transaction protocol {#apx:protocol}
+
+Our protocol for executing transactions is fairly straightforward: two-phase commit, and uses two-phase locking with retries to guarantee isolation.
+We employ a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
+However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
+
+To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
+During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
+an "increment" speculatively returns the incremented value).
+When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
+This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
+
+## Retwis workload {#apx:retwis}
+```{r}
+df <- subset(db("select * from tapir where stat_following_counts is not null and name like '%v0.14%'"),
+  nclients == 32
+  & initusers == 4096
+)
+df$grp <- with(df, sprintf("%s\n%s\nmix:%s/%s,\n%s", name, ccmode, mix, alpha, gen))
+
+histogram.facets <- function(df, measure, grp) {
+  d <- data.frame(x=c(),y=c(),version=c())
+  for (i in 1:nrow(df)) {
+    d <- rbind(d, df.histogram(df[i,measure], df[i,grp]))
+  }
+  return(d)
+}
+```
+
+```{r followers, fig.cap="CDF of the number of followers for users generated by the Kronecker synthetic graph generator, matching the power-law degree distribution of natural graphs.\\label{followers}"}
+d.follow <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_follower_counts', 'grp')
+ggplot(d.follow, aes(x=x, weight=y))+
+    stat_ecdf(color=c.blue)+
+    xlab('# followers / user (log scale)')+ylab('CDF (log scale)')+
+    scale_x_log10(breaks=c(1,10,100,1000))+scale_y_log10(breaks=c(0.1,0.5,1.0))+
+    theme_mine
+```
+
+```{r reposts, fig.cap="CDF of the number of times a post is reposted, matching traces from real workloads.
+Due to the power-law graph structure, if users tend to reposts popular recent posts on their timeline, it results in another power-law distribution.
+Note that that some posts are reposted to over a quarter of the graph (4000 total users).\\label{reposts}"}
+d.repost <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_repost_counts', 'grp')
+ggplot(d.repost, aes(x=x, weight=y))+
+    stat_ecdf(color=c.blue)+
+    xlab('# reposts')+ylab('count')+
+    scale_x_log10(breaks=c(1,10,100,1000))+scale_y_log10(breaks=c(0.1,0.2,0.4,0.6,0.8,1.0))+
+    xlab('# reposts (log scale)')+ylab('CDF (log scale)')+
+    theme_mine
+```
+
+With our Retwis workload, rather than approximating the skew in real-world workloads with a Zipfian as many other systems do, we attempt to simulate the behavior of social networks with a realistic synthetic graph and a simple model of user behavior for posting and reposting.
+
+For our synthetic graph, we use the Kronecker graph generator from the Graph 500 benchmark \cite{graph500}.
+This generator is designed to result in graphs with the same power-law degree distribution found in natural graphs.
+\autoref{followers} shows the cumulative distribution function (CDF) of the number of followers per user for the synthetic graph of approximately 4000 users, with an average number of followers of 16 (scale 12 with edgefactor of 16 in Graph500's terms) which we use in our performance results back in \autoref{eval}.
+Most users should have relatively few followers; we see that roughly 50% have fewer than 100 followers, while a very small number of users have over 1000 followers.
+
+We use a simple model of user behavior to determine when and which posts to repost.
+Each time we load the most recent posts in a timeline for a random user (uniformly selected), they are sorted by the number of times they have already been reposted, and a discrete geometric distribution, skewed toward 0, is used to select the number of these to repost.
+This results in the "viral" propagation effect that is observed in real social networks.
+\autoref{reposts} shows the distribution of the number of times a post was reposted, which is again a power-law distribution.
+Note that a small number of posts are reposted so much that they end up on over a quarter of users' timelines.

### 2015-02-14 rmd/paper.Rmd
some space hacking with figures

@@
 
-```{r stress, fig.cap="Throughput of raw Set operations., \\label{stress_tput}", fig.height=3.3}
+```{r stress, include=F}
 d <- data(db("
@@
 ```
+\begin{figure}[t]
+\centering
+\includegraphics{figure/stress-1.pdf}
+\caption{Throughput of raw Set operations.\label{stress_tput}}
+\end{figure}
 
@@
 
-```{r throughput, fig.cap="Throughput of social network workload (Retwis) with 4000 users.
-Leveraging commutativity prevents performance from falling over even when posts spread virally (repost-heavy).\\label{tput}"}
+```{r throughput, include=F}
 d <- data(db("
@@
 ```
+\begin{figure}[t]
+\centering
+\includegraphics{figure/throughput-1.pdf}
+\caption{Throughput of social network workload (Retwis) with 4000 users.
+Leveraging commutativity prevents performance from falling over even
+when posts spread virally (repost-heavy).\label{tput}}
+\end{figure}
 

### 2015-02-14 rmd/paper.Rmd
more space hacks

@@
 
-```{r followers, fig.cap="CDF of the number of followers for users generated by the Kronecker synthetic graph generator, matching the power-law degree distribution of natural graphs.\\label{followers}"}
+```{r followers, include=F}
 d.follow <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_follower_counts', 'grp')
@@
 ```
+\begin{figure}[t]\centering
+\includegraphics{figure/followers-1.pdf}
+\caption{CDF of the number of followers for users generated by the
+Kronecker synthetic graph generator, matching the power-law degree
+distribution of natural graphs.\label{followers}}
+\end{figure}
 
-```{r reposts, fig.cap="CDF of the number of times a post is reposted, matching traces from real workloads.
-Due to the power-law graph structure, if users tend to reposts popular recent posts on their timeline, it results in another power-law distribution.
-Note that that some posts are reposted to over a quarter of the graph (4000 total users).\\label{reposts}"}
+```{r reposts, include=F}
 d.repost <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_repost_counts', 'grp')
@@
 ```
+\begin{figure}[t]\centering
+\includegraphics{figure/reposts-1.pdf}
+\caption{CDF of the number of times a post is reposted, matching traces
+from real workloads. Due to the power-law graph structure, if users tend
+to reposts popular recent posts on their timeline, it results in another
+power-law distribution. Note that that some posts are reposted to over a
+quarter of the graph (4000 total users).\label{reposts}}
+\end{figure}
 

### 2015-02-14 rmd/paper.Rmd
fix references

@@
 If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model?
-What if the programmer could express the semantics they desire succintly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?
+What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?
 
@@
 This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control.
-Using insights from multi-core concurrent data structures, we show in Section \ref{sec:commutativity} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.
+Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.
 
@@
 # Commutativity {#comm}
-\label{sec:commutativity}
 
@@
 \begin{table}
+\resizebox{\columnwidth}{!}{%
 \centering
@@
 \end{tabular}
-\caption{\label{spec} Commutativity Specification for Set.}
+\caption{\label{spec} Commutativity Specification for Set.}}
 \end{table}
@@
 Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
-We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in Table \ref{tab:spec}.
+We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
 It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$.
@@
 
-For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in Table \ref{tab:spec}, but there may be many that express varying degrees of commutativity.
+For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity.
 The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute.
@@
 This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory.
-In Section \ref{sec:evaluation}, we show how it can be applied to distributed transactions.
+In \autoref{eval}, we show how it can be applied to distributed transactions.
 

### 2015-02-14 rmd/paper.Rmd
cutting with great abandon

@@
 Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
-
-Performance benefits come from understanding the properties of ADT operations: those that commute with each other can be performed concurrently, even on multiple copies of the record.
-This means that transactions whose operations commute abort less, approaching the performance without transactions.
-This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control.
-Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.
-
-Selecting the right data type for the job gives programmers a clean, precise way to express their desired behavior.
-For instance, rather than using a generic integer to generate unique identifiers, a `UUID` type, realizing that contiguity of ids isn't necessary, can be trivially parallelized and distributed.
-Though this is an extremely simple case (and nearly universally adopted optimization), it fits the same mold as more nuanced decisions, such as choosing to represent the number of times a given post has been "retweeted" as a `HyperLogLog`, which can efficiently yield the approximate number of unique users, rather than a costly precise `Set`.
-Though selecting data structures for the job at hand is nothing new to programmers, only a handful of databases, such as Redis, MongoDB or Riak, support this flexibility, and they do not use the abstraction it affords to enforce strongly consistent transactions.
+Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts.
+By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.
 
@@
 Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
-
-In the realm of eventual consistency, commutativity has been leveraged for convergence guarantees.
-RedBlue consistency allows executing commutative ("blue") operations locally, knowing they will eventually converge.
-Similarly, conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11} define commutative merge functions for all operations to ensure that replicas will converge.
-
-Lynx \cite{Zhang:SOSP13} uses knowledge of some commutative operations to make tracking serializability cheaper in chained transactions.
-Doppel \cite{Narula:OSDI14} added several explicitly commutative operations on records which they exploited to better handle common high-contention situations such as counters and "top-k lists" in the context of a single node multicore database.
-Finally, HyFlow \cite{Kim:EuroPar13}, a distributed transactional memory framework, reorders commutative operations on specific data types to execute before others to allow them to operate concurrently on a single version of a record.
+Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}.
+Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar} for reordering operations in the context of distributed transactional memory.
+We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.
 
@@
 If a pair of operations commute, then executing them in either order will produce the same result.
-Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
-We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
+Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
+We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
 It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$.
 In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.
-
-For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity.
-The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute.
-As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity.
-Data structure designers and users navigate these choices to optimize their programs.
+There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11} ADT designers can navigate these choices to optimize an implementation.
 
@@
 
-Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations, even without aborting, is a bottleneck which prevents scaling.
-The Ellen Degeneres selfie retweet is a prime example of this.
+Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations sequentially is a bottleneck which prevents scaling.
 In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity.
@@
 
-### Record splitting for parallelism
+**Record splitting.**
+Associativity, often paired with commutativity, can be leveraged to execute operations in parallel on a "split" record and the changes made to these splits can be merged to form a consistent whole.
+Doppel \cite{Narula:OSDI14} used this for a handful of operations.
+We propose generalizing this concept to ADTs: associative operations execute on replicas first, other operations wait to execute until the replicas have been brought back in sync.
 
-Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation.
-This allows them to *split* hot records onto multiple cores and execute commutative operations in parallel, *reconciling* them back to a single value before executing reads or non-commuting operations.
-This observation can be generalized via ADTs: operations can operate on copies of a record in parallel provided they commute with each other and provided they can be combined at the end of a phase before beginning to execute a new phase with a different set of operations that didn't commute with the first set.
-
-### Combining
-
-Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself.
-This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.
+**Combining.**
+Another spin on associative operations is to merge or *combine* operations as they come in.
+This is known as combining \cite{flatCombining,yew:combining-trees,funnels}, and it can drastically reduce contention.
+Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.
 
@@
 To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations.
-The transaction protocol employs standard two-phase commit, and two-phase locking, more details can be found in the appendix (\autoref{apx:protocol}).
+The transaction protocol employs standard two-phase commit, and two-phase locking; more details are in the appendix (\autoref{apx:protocol}).
 

### 2015-02-14 rmd/paper.Rmd
minor wording changes, fix urls in bibliography

@@
 In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong.
-Google, recognizing the benefit to programmer productivity, reintroduced serializable transactions to their next generation database, Spanner \cite{Spanner}.
-However, this leaves them with two extreme choices, with significant performance impact.
+Recognizing the benefit to correctness and productivity, some systems have reintroduced serializable transactions, such as Google's Spanner \cite{Spanner}.
+However, this leaves programmers with two extreme choices with significant performance impact.
 If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model?
@@
 
-Abstract data types (ADTs) are the solution to this problem.
+Abstract data types (ADTs) are the solution.
 Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
@@
 We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
-It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$.
-In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.
-There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11} ADT designers can navigate these choices to optimize an implementation.
+There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11}
 
@@
 If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record.
-This technique of raising the level of abstraction in transactions to operations is known as *transactional boosting* \cite{Herlihy:PPoPP08}.
-This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory.
-In \autoref{eval}, we show how it can be applied to distributed transactions.
+This technique is known as *transactional boosting* \cite{Herlihy:PPoPP08}.
+This straightforward use of commutativity was shown to significantly improve abort rates in software transactional memory.
+In \autoref{eval}, we show how we applied it to distributed transactions.
 
@@
 Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.
+
+# Data type selection
+
+Choosing the right data type for your application can drastically impact performance.
+Selecting the semantics that are most permissive or specialized for a particular use case gives the system the best chance of scaling performance.
+For a simple example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C \cite{TPCC}, which explicitly requires this, know well).
+Instead, a `UniqueID` type succinctly expresses that non-sequential IDs are okay, which can be implemented very efficiently.
+By allowing approximations or non-determinism, performance may be further improved.
+
+**Probabilistic data types** are a class of data types with *probabilistic* guarantees about their semantics, trading off some accuracy for better performance or storage.
+These include *bloom filters* \cite{bloom}, *hyperloglog* \cite{hyperloglog}, and *count-min sketch* \cite{countminsketch}.
+Hyperloglog, which also appears in Redis \cite{redis}, estimates the size of a set within a fixed error bound.
+Twitter's streaming analytics system \cite{summingbird} leverages these to handle their high data volume, and we can expect similar improvements.
+
+**Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type.
+This could allow copies to exist in different shards, asynchronously updating each other.
+By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state.
+Clients of these might find them more difficult to reason about, they might making this tradeoff in parts of the application where it otherwise cannot scale.
 
@@
 As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over.
-But with commutativity, we see that performance continues to scale even under this arduous load.
+But with commutativity, performance continues to scale even under this highly contentious load.
 

### 2015-02-17 rmd/paper.Rmd
cite ellen selfie
tighten up some wording

@@
     Out of the many NoSQL databases in use today, some that provide simple data structures for records, such as Redis and MongoDB, are now becoming popular.
-Building applications using these data structures rather than plain string values provides programmers with a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema.
-Currently, this expressiveness is used to ensure related values are co-located so they can be quickly accessed together (e.g.
-maps and other aggregates) and to provide complex atomic operations (e.g.
-set insertion).
-However, there are many more ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.
+Building applications out of these complex data types provides a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema.
+Currently this expressiveness is used in limited ways, such as providing atomic operations or ensuring related values are co-located.
+However, there are many ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.
 
-    In this work, we demonstrate several ways of leveraging data structure semantics in databases, focusing primarily on commutativity.
-Reasoning about operation reordering can allow transactions to execute concurrently that would conflict under traditional concurrency control.
+    In this work, we demonstrate several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity.
 Using Retwis, a Twitter clone built for Redis, as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks.
-We conclude that data types are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
+We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
 ---
@@
 The move to non-relational (NoSQL) databases was motivated by a desire for scalability and flexibility.
-People found that by not enforcing strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
-By forgoing the structure of relational schemas, they gained more direct control over performance at the cost of significant automatic optimization.
-The observation has been that relaxed or eventual consistency is a good fit for many applications built on top of these NoSQL because potential conflicts are difficult to observe in practice and that users are likely to accept minor inconsistencies, such as two tweets being out of temporal order on their Twitter timeline.
-However, this leaves much to chance; there is likely no guarantee that more significant inconsistencies are impossible.
+People found that by giving up strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
+In fact for many applications, relaxed consistency feels natural; because of inherent uncertainty in timing and connectivity, users are likely to accept minor inconsistencies such as two tweets being out of temporal order or needing to retry an action.
+However, this leaves much to chance: there is likely no guarantee that more significant inconsistencies are impossible.
 
-In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong.
-Recognizing the benefit to correctness and productivity, some systems have reintroduced serializable transactions, such as Google's Spanner \cite{Spanner}.
-However, this leaves programmers with two extreme choices with significant performance impact.
-If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model?
-What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?
+When consistency truly is critical, developers can enforce stronger guarantees manually, or use serializable transactions in systems like Google's Spanner [@Spanner], but this leaves them with two extreme choices with a significant performance gap.
+If certain parts of an application can tolerate imprecision, why not capture those properties in the programming model?
+Is there a way programmers can express the semantics they desire succinctly and precisely, helping the database optimize performance and scalability, without sacrificing flexibility?
 
-Abstract data types (ADTs) are the solution.
-Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
-Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts.
-By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.
+We propose abstract data types (ADTs) as the solution.
+Rather than limiting the records in databases to primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
+We demonstrate one way of using commutativity of ADT operations to reduce transaction aborts and explore several other ways of leveraging commutativity and data types to improve database performance and allow programmers to make tradeoffs between performance and precision.
 
@@
 Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}.
-Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar} for reordering operations in the context of distributed transactional memory.
+Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar13} for reordering operations in the context of distributed transactional memory.
 We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.
@@
 By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state.
-Clients of these might find them more difficult to reason about, they might making this tradeoff in parts of the application where it otherwise cannot scale.
+Clients of these might find them more difficult to reason about but may consider making this tradeoff in parts of the application where it otherwise cannot scale.
 
@@
 However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
-As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over.
+As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record \cite{ellenselfie}, so too does our baseline fall over.
 But with commutativity, performance continues to scale even under this highly contentious load.

### 2015-02-17 rmd/paper.Rmd
smaller figures, convert to Pandoc citations, more wordsmithing

@@
 ```{r setup, include=FALSE}
-opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.2, fig.height=3)
+opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.6, fig.height=3)
 ```
@@
 People found that by giving up strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
-In fact for many applications, relaxed consistency feels natural; because of inherent uncertainty in timing and connectivity, users are likely to accept minor inconsistencies such as two tweets being out of temporal order or needing to retry an action.
-However, this leaves much to chance: there is likely no guarantee that more significant inconsistencies are impossible.
-
-When consistency truly is critical, developers can enforce stronger guarantees manually, or use serializable transactions in systems like Google's Spanner [@Spanner], but this leaves them with two extreme choices with a significant performance gap.
+Because of inherent uncertainty in timing and connectivity, in many cases users are likely to accept minor inconsistencies such as two tweets being out of temporal order or needing to retry an action.
+In such cases, relaxed consistency feels like a natural solution, but it leaves much to chance: there is likely no guarantee that more significant inconsistencies are impossible.
+When consistency is critical, developers can enforce stronger guarantees manually, or use serializable transactions in systems like Google's Spanner [@Spanner], but this leaves them with two extremes with a significant performance gap.
 If certain parts of an application can tolerate imprecision, why not capture those properties in the programming model?
@@
 Rather than limiting the records in databases to primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
-We demonstrate one way of using commutativity of ADT operations to reduce transaction aborts and explore several other ways of leveraging commutativity and data types to improve database performance and allow programmers to make tradeoffs between performance and precision.
+In this work we explore several ways of leveraging commutativity and data types to improve database performance and allow programmers to make tradeoffs between performance and precision, starting by demonstrating one way of using commutativity to reduce transaction aborts.
 
@@
 Commutativity is well known, especially in distributed systems, for enabling important optimizations.
-Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict.
+Since the 80s, commutativity has been exploited by database systems designers [@Weihl:1988,Fekete:90], within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict.
 Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
-Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}.
-Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar13} for reordering operations in the context of distributed transactional memory.
+Eventually consistent databases use commutativity for convergence in work such as RedBlue consistency [@Li:OSDI12] and conflict-free replicated data types (CRDTs) [@Shapiro:SSS11].
+Other systems specialize for commutative operations to improve transaction processing, such as Lynx [@Zhang:SOSP13] for tracking serializability, and Doppel [@Narula:OSDI14] for executing operations in parallel on highly contended records, and HyFlow [@Kim:EuroPar13] for reordering operations in the context of distributed transactional memory.
 We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.
-
-## Commutativity Specifications
 
@@
 If a pair of operations commute, then executing them in either order will produce the same result.
-Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
+Using the definitions from [@Kulkarni:PLDI11], whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
 We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
-There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11}
+There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.[@Kulkarni:PLDI11]
 
-## Transactional Boosting
 If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record.
-This technique is known as *transactional boosting* \cite{Herlihy:PPoPP08}.
+This technique is known as *transactional boosting* [@Herlihy:PPoPP08].
 This straightforward use of commutativity was shown to significantly improve abort rates in software transactional memory.
 In \autoref{eval}, we show how we applied it to distributed transactions.
-
-## Other opportunities
-
-Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations sequentially is a bottleneck which prevents scaling.
-In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity.
-We mention just a few here which we have yet to evaluate.
-
-**Record splitting.**
-Associativity, often paired with commutativity, can be leveraged to execute operations in parallel on a "split" record and the changes made to these splits can be merged to form a consistent whole.
-Doppel \cite{Narula:OSDI14} used this for a handful of operations.
-We propose generalizing this concept to ADTs: associative operations execute on replicas first, other operations wait to execute until the replicas have been brought back in sync.
-
-**Combining.**
-Another spin on associative operations is to merge or *combine* operations as they come in.
-This is known as combining \cite{flatCombining,yew:combining-trees,funnels}, and it can drastically reduce contention.
-Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.
 
@@
 Selecting the semantics that are most permissive or specialized for a particular use case gives the system the best chance of scaling performance.
-For a simple example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C \cite{TPCC}, which explicitly requires this, know well).
+For a simple example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C [@TPCC], which explicitly requires this, know well).
 Instead, a `UniqueID` type succinctly expresses that non-sequential IDs are okay, which can be implemented very efficiently.
@@
 **Probabilistic data types** are a class of data types with *probabilistic* guarantees about their semantics, trading off some accuracy for better performance or storage.
-These include *bloom filters* \cite{bloom}, *hyperloglog* \cite{hyperloglog}, and *count-min sketch* \cite{countminsketch}.
-Hyperloglog, which also appears in Redis \cite{redis}, estimates the size of a set within a fixed error bound.
-Twitter's streaming analytics system \cite{summingbird} leverages these to handle their high data volume, and we can expect similar improvements.
+These include *bloom filters* [@bloom], *hyperloglog* [@hyperloglog], and *count-min sketch* [@countminsketch].
+Hyperloglog, which also appears in Redis [@redis], estimates the size of a set within a fixed error bound.
+Twitter's streaming analytics system [@summingbird] leverages these to handle their high data volume, and we can expect similar improvements.
 
@@
 Commutativity comes into play in the locking scheme.
-Using the algorithms from \cite{Kulkarni:PLDI11} and our commutativity specifications, we design an abstract lock for each record type.
+Using the algorithms from [@Kulkarni:PLDI11] and our commutativity specifications, we design an abstract lock for each record type.
 Our `SortedSet`, for instance, has an `add` mode which allows all insertion operations to commute, but disallows operations like `contains` or `size`.
@@
 
-```{r stress, include=F}
+```{r stress, include=F, fig.height=2.4}
 d <- data(db("
@@
 ggplot(d.u, aes(x=nclients, y=throughput/1000, group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=max, geom="line")+
+    stat_summary(fun.y=max, geom="line", size=0.4)+
     xlab('Concurrent clients')+ylab('Throughput (k/sec)')+
@@
     facet_grid(dist~opmix)+
-    theme_mine+theme(legend.position='bottom', legend.direction='horizontal', legend.title.align=1)+
+    theme_mine+theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 
-```{r throughput, include=F}
+```{r throughput, include=F, fig.height=2.0}
 d <- data(db("
@@
 d.u <- subset(d, nshards == 4 & initusers == 4096 & grepl('geom_repost|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput, group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=mean, geom="line")+
-    xlab('Concurrent clients')+ylab('Throughput (transactions / sec)')+
+ggplot(d.u, aes(x=nclients, y=throughput/1000, group=cc, fill=cc, color=cc, linetype=cc))+
+    stat_summary(fun.y=mean, geom="line", size=0.4)+
+    xlab('Concurrent clients')+ylab('Throughput (k trans. / sec)')+
     expand_limits(y=0)+
     facet_wrap(~workload)+
-    theme_mine+theme(legend.position='bottom', legend.direction='horizontal', legend.title.align=1)+
+    theme_mine+theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
-As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record \cite{ellenselfie}, so too does our baseline fall over.
+As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record [@ellenselfie], so too does our baseline fall over.
 But with commutativity, performance continues to scale even under this highly contentious load.
@@
 # Appendix
-
 ## Transaction protocol {#apx:protocol}
@@
 This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
+
+##  Other opportunities for commutativity
+
+Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations sequentially is a bottleneck which prevents scaling.
+In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity.
+We mention just a few here which we have yet to evaluate.
+
+**Record splitting.**
+Associativity, often paired with commutativity, can be leveraged to execute operations in parallel on a "split" record and the changes made to these splits can be merged to form a consistent whole.
+Doppel [@Narula:OSDI14] used this for a handful of operations.
+We propose generalizing this concept to ADTs: associative operations execute on replicas first, other operations wait to execute until the replicas have been brought back in sync.
+
+**Combining.**
+Another spin on associative operations is to merge or *combine* operations as they come in.
+This is known as combining [@flatCombining,yew:combining-trees,funnels], and it can drastically reduce contention.
+Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.
 
@@
 
-For our synthetic graph, we use the Kronecker graph generator from the Graph 500 benchmark \cite{graph500}.
+For our synthetic graph, we use the Kronecker graph generator from the Graph 500 benchmark [@graph500].
 This generator is designed to result in graphs with the same power-law degree distribution found in natural graphs.

### 2015-02-17 rmd/paper.Rmd
reorder some stuff

@@
 d$dist <- factor(revalue(d$alpha, c('0.6'='Zipf: 0.6', '-1'='Uniform')))
-d.u <- subset(d, nshards == 4 & nkeys == 10000 & (alpha == '0.6' | alpha == '-1') & grepl('update_heavy|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput/1000, group=cc, fill=cc, color=cc, linetype=cc))+
+d.u <- subset(d, nshards == 4 & nkeys == 10000 
+    & (alpha == '0.6' | alpha == '-1') 
+    & grepl('update_heavy|read_heavy', mix)
+)
+ggplot(d.u, aes(x=nclients, y=throughput/1000,
+        group=cc, fill=cc, color=cc, linetype=cc))+
     stat_summary(fun.y=max, geom="line", size=0.4)+
@@
     facet_grid(dist~opmix)+
-    theme_mine+theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
+    theme_mine+
+    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 ))
-d.u <- subset(d, nshards == 4 & initusers == 4096 & grepl('geom_repost|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput/1000, group=cc, fill=cc, color=cc, linetype=cc))+
+d.u <- subset(d, nshards == 4 & initusers == 4096 
+                & grepl('geom_repost|read_heavy', mix))
+ggplot(d.u, aes(x=nclients, y=throughput/1000, 
+        group=cc, fill=cc, color=cc, linetype=cc))+
     stat_summary(fun.y=mean, geom="line", size=0.4)+
@@
     facet_wrap(~workload)+
-    theme_mine+theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
+    theme_mine+
+    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 # Appendix
-## Transaction protocol {#apx:protocol}
 
-Our protocol for executing transactions is fairly straightforward: two-phase commit, and uses two-phase locking with retries to guarantee isolation.
-We employ a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
-However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
-
-To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
-During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
-an "increment" speculatively returns the incremented value).
-When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
-This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
+Many ideas and details did not fit into the core of the paper, but because they may help interpret our findings and give more of a sense of future directions, we include them here for anyone interested.
 
@@
 
+## Transaction protocol {#apx:protocol}
+
+Our protocol for executing transactions is fairly straightforward: two-phase commit, and uses two-phase locking with retries to guarantee isolation.
+We employ a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
+However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
+
+To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
+During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
+an "increment" speculatively returns the incremented value).
+When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
+This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
+
 ## Retwis workload {#apx:retwis}
 ```{r}
-df <- subset(db("select * from tapir where stat_following_counts is not null and name like '%v0.14%'"),
+df <- subset(db("
+    select * from tapir where stat_following_counts is not null
+    and name like '%v0.14%'
+"),
   nclients == 32
@@
 ```{r followers, include=F}
-d.follow <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_follower_counts', 'grp')
+d.follow <- histogram.facets(subset(df,
+    initusers == 4096 & mix == 'geom_repost'
+), 'stat_follower_counts', 'grp')
 ggplot(d.follow, aes(x=x, weight=y))+
@@
 ```{r reposts, include=F}
-d.repost <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_repost_counts', 'grp')
+d.repost <- histogram.facets(
+    subset(df, initusers == 4096 & mix == 'geom_repost')
+, 'stat_repost_counts', 'grp')
 ggplot(d.repost, aes(x=x, weight=y))+
@@
     xlab('# reposts')+ylab('count')+
-    scale_x_log10(breaks=c(1,10,100,1000))+scale_y_log10(breaks=c(0.1,0.2,0.4,0.6,0.8,1.0))+
+    scale_x_log10(breaks=c(1,10,100,1000))+
+    scale_y_log10(breaks=c(0.1,0.2,0.4,0.6,0.8,1.0))+
     xlab('# reposts (log scale)')+ylab('CDF (log scale)')+

### 2015-02-17 rmd/paper.Rmd
wordsmithing

@@
 
-    In this work, we demonstrate several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity.
-Using Retwis, a Twitter clone built for Redis, as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks.
+  In this work, we explore several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity.
+Using a Twitter clone as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks.
 We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
@@
 Hyperloglog, which also appears in Redis [@redis], estimates the size of a set within a fixed error bound.
-Twitter's streaming analytics system [@summingbird] leverages these to handle their high data volume, and we can expect similar improvements.
+Twitter's streaming analytics system [@summingbird] leverages these to handle their high data volume, and we expect similar benefit.
 
@@
 By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state.
-Clients of these might find them more difficult to reason about but may consider making this tradeoff in parts of the application where it otherwise cannot scale.
+Clients may find them more difficult to reason about but might make this tradeoff in parts of the application where it otherwise cannot scale.
 

### 2015-02-17 rmd/paper.Rmd
reintroduce combining/splitting into body, other tweaks

@@
 Building applications out of these complex data types provides a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema.
-Currently this expressiveness is used in limited ways, such as providing atomic operations or ensuring related values are co-located.
-However, there are many ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.
+Currently this capability is leveraged in limited ways, such as to ensure related values are co-located, or for atomic updates.
+There are many ways data types can be used to make databases more efficient that are not yet being exploited.
 
-  In this work, we explore several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity.
+  We explore several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity.
 Using a Twitter clone as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks.
-We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
+We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, making databases more safe and scalable.
 ---
@@
 Commutativity is well known, especially in distributed systems, for enabling important optimizations.
-Since the 80s, commutativity has been exploited by database systems designers [@Weihl:1988,Fekete:90], within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict.
+Since the 80s, commutativity has been exploited by database systems designers [@Weihl:1988;@Fekete:90] within the safe confines of relational models, where complete control of the data structures allows systems to determine when transactions may conflict.
 Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
 Eventually consistent databases use commutativity for convergence in work such as RedBlue consistency [@Li:OSDI12] and conflict-free replicated data types (CRDTs) [@Shapiro:SSS11].
-Other systems specialize for commutative operations to improve transaction processing, such as Lynx [@Zhang:SOSP13] for tracking serializability, and Doppel [@Narula:OSDI14] for executing operations in parallel on highly contended records, and HyFlow [@Kim:EuroPar13] for reordering operations in the context of distributed transactional memory.
-We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.
+Other systems specialize for commutative operations to improve transaction processing, such as Lynx [@Zhang:SOSP13] for tracking serializability, Doppel [@Narula:OSDI14] for executing operations in parallel on highly contended records, and HyFlow [@Kim:EuroPar13] for reordering operations in the context of distributed transactional memory.
+We propose unifying and generalizing these under the abstraction afforded by ADTs.
 
 \begin{table}
+\centering
 \resizebox{\columnwidth}{!}{%
-\centering
 \begin{tabular}{lll}
@@
 We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
-There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.[@Kulkarni:PLDI11]
+There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.
 
-If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record.
+**Transaction boosting.** If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record.
 This technique is known as *transactional boosting* [@Herlihy:PPoPP08].
@@
 
+**Combining.** Associativity, often paired with commutativity, allows some concurrent operations to be *combined* before being applied to the data structure itself.
+*Combining* [@flatCombining;@yew:combining-trees;@funnels] can drastically reduce contention on shared data structures.
+This technique could be applied to hot records, similar to *splitting* in Doppel [@Narula:OSDI14], to avoid bottlenecking on a single shard.
+
 # Data type selection
 
-Choosing the right data type for your application can drastically impact performance.
-Selecting the semantics that are most permissive or specialized for a particular use case gives the system the best chance of scaling performance.
-For a simple example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C [@TPCC], which explicitly requires this, know well).
+Choosing an ADT with semantics specialized for a particular use case gives the system the best chance of scaling performance.
+For example, rather than using a counter, which must return the next number in the sequence (which is difficult to scale, as users of TPC-C [@TPCC] know well)
+For example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C [@TPCC], which explicitly requires this, know well).
 Instead, a `UniqueID` type succinctly expresses that non-sequential IDs are okay, which can be implemented very efficiently.
@@
 
-**Probabilistic data types** are a class of data types with *probabilistic* guarantees about their semantics, trading off some accuracy for better performance or storage.
-These include *bloom filters* [@bloom], *hyperloglog* [@hyperloglog], and *count-min sketch* [@countminsketch].
-Hyperloglog, which also appears in Redis [@redis], estimates the size of a set within a fixed error bound.
-Twitter's streaming analytics system [@summingbird] leverages these to handle their high data volume, and we expect similar benefit.
+**Probabilistic data types** such as *bloom filters* [@bloom], *hyperloglogs* [@hyperloglog], and *count-min sketches* [@countminsketch] trade off accuracy (within fixed bounds) for better performance or storage.
+Twitter's streaming analytics system [@summingbird] and many machine learning algorithms leverage these to handle high data volume, and we expect similar benefit.
 
 **Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type.
-This could allow copies to exist in different shards, asynchronously updating each other.
-By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state.
-Clients may find them more difficult to reason about but might make this tradeoff in parts of the application where it otherwise cannot scale.
+Copies of a record could exist in different shards, asynchronously updating each other.
+By defining the same kind of *merge* function as traditional CRDTs, these copies could ensure they all converge to the same state.
+Clients may find them more difficult to reason about but might make that tradeoff in parts of the application where it otherwise cannot scale.
 
@@
 
-```{r throughput, include=F, fig.height=2.0}
+```{r throughput, include=F, fig.height=2.1}
 d <- data(db("
@@
 **Case study: Retwis.** 
-To understand performance of more typical of web workloads, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis.
+To understand performance on a typical web workload, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis [@redis].
 We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
-On top of Retwis's basic functionality, we also add a "repost" option that works like Twitter's "retweet".
+On top of Retwis's basic functionality, we also add a "repost" action that behaves like Twitter's "retweet".
 We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
@@
 
-Many ideas and details did not fit into the core of the paper, but because they may help interpret our findings and give more of a sense of future directions, we include them here for anyone interested.
-
-##  Other opportunities for commutativity
-
-Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations sequentially is a bottleneck which prevents scaling.
-In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity.
-We mention just a few here which we have yet to evaluate.
-
-**Record splitting.**
-Associativity, often paired with commutativity, can be leveraged to execute operations in parallel on a "split" record and the changes made to these splits can be merged to form a consistent whole.
-Doppel [@Narula:OSDI14] used this for a handful of operations.
-We propose generalizing this concept to ADTs: associative operations execute on replicas first, other operations wait to execute until the replicas have been brought back in sync.
-
-**Combining.**
-Another spin on associative operations is to merge or *combine* operations as they come in.
-This is known as combining [@flatCombining,yew:combining-trees,funnels], and it can drastically reduce contention.
-Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.
+Some details did not fit into the core of the paper, but because they may help interpret our findings and give a sense of future directions, we include them here for anyone interested.
 
@@
 
-```{r followers, include=F}
+```{r followers, include=F, fig.width=2.5, fig.height=2.5}
 d.follow <- histogram.facets(subset(df,
@@
 
-```{r reposts, include=F}
+```{r reposts, include=F, fig.width=2.5, fig.height=2.5}
 d.repost <- histogram.facets(

### 2015-03-08 poster/abstract.tex
add poster abstract

--- 2015-02-17 rmd/paper.Rmd
@@
----
-title: 'Claret: Using Data Types for Highly Concurrent Distributed Transactions'
-preprint: true
+\documentclass[10pt]{article}
+\usepackage[scaled=0.8]{inconsolata}
 
-author:
-  - {family: Holt,  given: Brandon, affiliation: 1, email: bholt}
-  - {family: Zhang, given: Irene,   affiliation: 1, email: iyzhang}
-  - {family: Ports, given: Dan,     affiliation: 1, email: dkp}
-  - {family: Oskin, given: Mark,    affiliation: 1, email: oskin}
-  - {family: Ceze,  given: Luis,    affiliation: 1, email: luisceze}
+\begin{document}
 
-organization:
-  - {id: 1, name: University of Washington}
+  \title{Claret: Using Data Types for Highly Concurrent Distributed Transactions}
+  \author{
+    Brandon Holt, Irene Zhang, Dan Ports, Mark Oskin, Luis Ceze \\
+    University of Washington \\
+    {\tt \{bholt,iyzhang,drkp,oskin,luisceze\}@cs.washington.edu}
+  }
 
-conference:
-  name: PaPoC 2015
-  location: 'April 21, 2015, Bordeaux, France'
-  year: 2015
+  \maketitle
 
-doi: 0
-
-layout: sigplanconf
-bibliography: biblio.bib
-output:
-  pdf_document:
-    fig_caption: yes
-
-abstract: |
+  \begin{abstract}
     Out of the many NoSQL databases in use today, some that provide simple data structures for records, such as Redis and MongoDB, are now becoming popular.
@@
 We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, making databases more safe and scalable.
----
-```{r setup, include=FALSE}
-opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.6, fig.height=3)
-```
+  \end{abstract}
 
-# Introduction
-
-The move to non-relational (NoSQL) databases was motivated by a desire for scalability and flexibility.
-People found that by giving up strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
-Because of inherent uncertainty in timing and connectivity, in many cases users are likely to accept minor inconsistencies such as two tweets being out of temporal order or needing to retry an action.
-In such cases, relaxed consistency feels like a natural solution, but it leaves much to chance: there is likely no guarantee that more significant inconsistencies are impossible.
-When consistency is critical, developers can enforce stronger guarantees manually, or use serializable transactions in systems like Google's Spanner [@Spanner], but this leaves them with two extremes with a significant performance gap.
-If certain parts of an application can tolerate imprecision, why not capture those properties in the programming model?
-Is there a way programmers can express the semantics they desire succinctly and precisely, helping the database optimize performance and scalability, without sacrificing flexibility?
-
-We propose abstract data types (ADTs) as the solution.
-Rather than limiting the records in databases to primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.
-In this work we explore several ways of leveraging commutativity and data types to improve database performance and allow programmers to make tradeoffs between performance and precision, starting by demonstrating one way of using commutativity to reduce transaction aborts.
-
-# Commutativity {#comm}
-
-Commutativity is well known, especially in distributed systems, for enabling important optimizations.
-Since the 80s, commutativity has been exploited by database systems designers [@Weihl:1988;@Fekete:90] within the safe confines of relational models, where complete control of the data structures allows systems to determine when transactions may conflict.
-Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
-Eventually consistent databases use commutativity for convergence in work such as RedBlue consistency [@Li:OSDI12] and conflict-free replicated data types (CRDTs) [@Shapiro:SSS11].
-Other systems specialize for commutative operations to improve transaction processing, such as Lynx [@Zhang:SOSP13] for tracking serializability, Doppel [@Narula:OSDI14] for executing operations in parallel on highly contended records, and HyFlow [@Kim:EuroPar13] for reordering operations in the context of distributed transactional memory.
-We propose unifying and generalizing these under the abstraction afforded by ADTs.
-
-\begin{table}
-\centering
-\resizebox{\columnwidth}{!}{%
-\begin{tabular}{lll}
-\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
-\hline
-\texttt{add(x): void} & \texttt{add(y)} & $\forall x, y$ \\
-\texttt{remove(x): void} & \texttt{remove(y)} & $\forall x, y$ \\
-    & \texttt{add(y)} & $x \ne y$ \\
-\texttt{size(): int} & \texttt{add(x)} & $x \in Set$ \\
-    & \texttt{remove(x)} & $x \notin Set$ \\
-\texttt{contains(x): bool} & \texttt{add(y)} & $x \ne y \lor y \in Set$ \\
-    & \texttt{remove(y)} & $x \ne y \lor y \notin Set$ \\
-    & \texttt{size()} & $\forall x$ \\
-    \hline
-\end{tabular}
-\caption{\label{spec} Commutativity Specification for Set.}}
-\end{table}
-
-Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced.
-If a pair of operations commute, then executing them in either order will produce the same result.
-Using the definitions from [@Kulkarni:PLDI11], whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target.
-We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}.
-There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.
-
-**Transaction boosting.** If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record.
-This technique is known as *transactional boosting* [@Herlihy:PPoPP08].
-This straightforward use of commutativity was shown to significantly improve abort rates in software transactional memory.
-In \autoref{eval}, we show how we applied it to distributed transactions.
-
-**Combining.** Associativity, often paired with commutativity, allows some concurrent operations to be *combined* before being applied to the data structure itself.
-*Combining* [@flatCombining;@yew:combining-trees;@funnels] can drastically reduce contention on shared data structures.
-This technique could be applied to hot records, similar to *splitting* in Doppel [@Narula:OSDI14], to avoid bottlenecking on a single shard.
-
-# Data type selection
-
-Choosing an ADT with semantics specialized for a particular use case gives the system the best chance of scaling performance.
-For example, rather than using a counter, which must return the next number in the sequence (which is difficult to scale, as users of TPC-C [@TPCC] know well)
-For example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C [@TPCC], which explicitly requires this, know well).
-Instead, a `UniqueID` type succinctly expresses that non-sequential IDs are okay, which can be implemented very efficiently.
-By allowing approximations or non-determinism, performance may be further improved.
-
-**Probabilistic data types** such as *bloom filters* [@bloom], *hyperloglogs* [@hyperloglog], and *count-min sketches* [@countminsketch] trade off accuracy (within fixed bounds) for better performance or storage.
-Twitter's streaming analytics system [@summingbird] and many machine learning algorithms leverage these to handle high data volume, and we expect similar benefit.
-
-**Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type.
-Copies of a record could exist in different shards, asynchronously updating each other.
-By defining the same kind of *merge* function as traditional CRDTs, these copies could ensure they all converge to the same state.
-Clients may find them more difficult to reason about but might make that tradeoff in parts of the application where it otherwise cannot scale.
-
-# Evaluation {#eval}
-To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations.
-The transaction protocol employs standard two-phase commit, and two-phase locking; more details are in the appendix (\autoref{apx:protocol}).
-
-Commutativity comes into play in the locking scheme.
-Using the algorithms from [@Kulkarni:PLDI11] and our commutativity specifications, we design an abstract lock for each record type.
-Our `SortedSet`, for instance, has an `add` mode which allows all insertion operations to commute, but disallows operations like `contains` or `size`.
-As a baseline, we implement a standard reader/writer locking scheme that allows all read-only operations to execute concurrently, but enforces that only one transaction may modify a record at a time.
-
-These experiments were run with 4 shards on 4 local nodes, each with 8-core 2GHz Xeon E5335 processors.
-
-```{r stress, include=F, fig.height=2.4}
-d <- data(db("
-    select * from tapir where 
-    total_time is not null
-    and name like 'stress-v0.14%'
-    ", factors=c('nshards', 'nclients'), numeric=c('total_time', 'txn_count')))
-d$opmix <- factor(revalue(d$mix, c(
-    'mostly_update'='35% read\n65% update',
-    'update_heavy'='50% read\n50% update',
-    'read_heavy'='90% read\n10% update'
-)))
-d$dist <- factor(revalue(d$alpha, c('0.6'='Zipf: 0.6', '-1'='Uniform')))
-d.u <- subset(d, nshards == 4 & nkeys == 10000 
-    & (alpha == '0.6' | alpha == '-1') 
-    & grepl('update_heavy|read_heavy', mix)
-)
-ggplot(d.u, aes(x=nclients, y=throughput/1000,
-        group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=max, geom="line", size=0.4)+
-    xlab('Concurrent clients')+ylab('Throughput (k/sec)')+
-    expand_limits(y=0)+
-    facet_grid(dist~opmix)+
-    theme_mine+
-    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
-    cc_scales(title='Concurrency\ncontrol:')
-```
-\begin{figure}[t]
-\centering
-\includegraphics{figure/stress-1.pdf}
-\caption{Throughput of raw Set operations.\label{stress_tput}}
-\end{figure}
-
-**Microbenchmark: Set operations.** We first evaluate performance with a simple workload consisting of a raw mix of `Set` operations randomly distributed over 10,000 keys.
-We use both a uniform random distribution as well as a skewed Zipfian distribution with a coefficient of 0.6.
-In \autoref{stress_tput}, we see that commutative transactions perform strictly better, showing the most pronounced benefit over the more update-heavy, skewed workload.
-
-```{r throughput, include=F, fig.height=2.1}
-d <- data(db("
-    select * from tapir where 
-    generator_time is not null and total_time is not null
-    and name like 'claret-v0.14%'
-  ",
-    factors=c('nshards', 'nclients'),
-    numeric=c('total_time', 'txn_count')
-))
-d.u <- subset(d, nshards == 4 & initusers == 4096 
-                & grepl('geom_repost|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput/1000, 
-        group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=mean, geom="line", size=0.4)+
-    xlab('Concurrent clients')+ylab('Throughput (k trans. / sec)')+
-    expand_limits(y=0)+
-    facet_wrap(~workload)+
-    theme_mine+
-    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
-    cc_scales(title='Concurrency\ncontrol:')
-```
-\begin{figure}[t]
-\centering
-\includegraphics{figure/throughput-1.pdf}
-\caption{Throughput of social network workload (Retwis) with 4000 users.
-Leveraging commutativity prevents performance from falling over even
-when posts spread virally (repost-heavy).\label{tput}}
-\end{figure}
-
-**Case study: Retwis.** 
-To understand performance on a typical web workload, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis [@redis].
-We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
-On top of Retwis's basic functionality, we also add a "repost" action that behaves like Twitter's "retweet".
-We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
-More about this workload is in the appendix (\autoref{apx:retwis}).
-
-\autoref{tput} shows the results of this simulation.
-When most of the traffic is content consumption (reading timelines), both systems perform well enough.
-However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
-As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record [@ellenselfie], so too does our baseline fall over.
-But with commutativity, performance continues to scale even under this highly contentious load.
-
-# Appendix
-
-Some details did not fit into the core of the paper, but because they may help interpret our findings and give a sense of future directions, we include them here for anyone interested.
-
-## Transaction protocol {#apx:protocol}
-
-Our protocol for executing transactions is fairly straightforward: two-phase commit, and uses two-phase locking with retries to guarantee isolation.
-We employ a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
-However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
-
-To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
-During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
-an "increment" speculatively returns the incremented value).
-When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
-This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
-
-## Retwis workload {#apx:retwis}
-```{r}
-df <- subset(db("
-    select * from tapir where stat_following_counts is not null
-    and name like '%v0.14%'
-"),
-  nclients == 32
-  & initusers == 4096
-)
-df$grp <- with(df, sprintf("%s\n%s\nmix:%s/%s,\n%s", name, ccmode, mix, alpha, gen))
-
-histogram.facets <- function(df, measure, grp) {
-  d <- data.frame(x=c(),y=c(),version=c())
-  for (i in 1:nrow(df)) {
-    d <- rbind(d, df.histogram(df[i,measure], df[i,grp]))
-  }
-  return(d)
-}
-```
-
-```{r followers, include=F, fig.width=2.5, fig.height=2.5}
-d.follow <- histogram.facets(subset(df,
-    initusers == 4096 & mix == 'geom_repost'
-), 'stat_follower_counts', 'grp')
-ggplot(d.follow, aes(x=x, weight=y))+
-    stat_ecdf(color=c.blue)+
-    xlab('# followers / user (log scale)')+ylab('CDF (log scale)')+
-    scale_x_log10(breaks=c(1,10,100,1000))+scale_y_log10(breaks=c(0.1,0.5,1.0))+
-    theme_mine
-```
-\begin{figure}[t]\centering
-\includegraphics{figure/followers-1.pdf}
-\caption{CDF of the number of followers for users generated by the
-Kronecker synthetic graph generator, matching the power-law degree
-distribution of natural graphs.\label{followers}}
-\end{figure}
-
-```{r reposts, include=F, fig.width=2.5, fig.height=2.5}
-d.repost <- histogram.facets(
-    subset(df, initusers == 4096 & mix == 'geom_repost')
-, 'stat_repost_counts', 'grp')
-ggplot(d.repost, aes(x=x, weight=y))+
-    stat_ecdf(color=c.blue)+
-    xlab('# reposts')+ylab('count')+
-    scale_x_log10(breaks=c(1,10,100,1000))+
-    scale_y_log10(breaks=c(0.1,0.2,0.4,0.6,0.8,1.0))+
-    xlab('# reposts (log scale)')+ylab('CDF (log scale)')+
-    theme_mine
-```
-\begin{figure}[t]\centering
-\includegraphics{figure/reposts-1.pdf}
-\caption{CDF of the number of times a post is reposted, matching traces
-from real workloads. Due to the power-law graph structure, if users tend
-to reposts popular recent posts on their timeline, it results in another
-power-law distribution. Note that that some posts are reposted to over a
-quarter of the graph (4000 total users).\label{reposts}}
-\end{figure}
-
-With our Retwis workload, rather than approximating the skew in real-world workloads with a Zipfian as many other systems do, we attempt to simulate the behavior of social networks with a realistic synthetic graph and a simple model of user behavior for posting and reposting.
-
-For our synthetic graph, we use the Kronecker graph generator from the Graph 500 benchmark [@graph500].
-This generator is designed to result in graphs with the same power-law degree distribution found in natural graphs.
-\autoref{followers} shows the cumulative distribution function (CDF) of the number of followers per user for the synthetic graph of approximately 4000 users, with an average number of followers of 16 (scale 12 with edgefactor of 16 in Graph500's terms) which we use in our performance results back in \autoref{eval}.
-Most users should have relatively few followers; we see that roughly 50% have fewer than 100 followers, while a very small number of users have over 1000 followers.
-
-We use a simple model of user behavior to determine when and which posts to repost.
-Each time we load the most recent posts in a timeline for a random user (uniformly selected), they are sorted by the number of times they have already been reposted, and a discrete geometric distribution, skewed toward 0, is used to select the number of these to repost.
-This results in the "viral" propagation effect that is observed in real social networks.
-\autoref{reposts} shows the distribution of the number of times a post was reposted, which is again a power-law distribution.
-Note that a small number of posts are reposted so much that they end up on over a quarter of users' timelines.
+\end{document}

### 2015-03-08 rmd/paper.Rmd
fix Dan's email (oops)

@@
   - {family: Zhang, given: Irene,   affiliation: 1, email: iyzhang}
-  - {family: Ports, given: Dan,     affiliation: 1, email: dkp}
+  - {family: Ports, given: Dan,     affiliation: 1, email: drkp}
   - {family: Oskin, given: Mark,    affiliation: 1, email: oskin}

### 2015-03-14 rmd/paper.Rmd
add copyright & conference info, doi

@@
 title: 'Claret: Using Data Types for Highly Concurrent Distributed Transactions'
-preprint: true
+documentclass:
+  name: sigplanconf
+  options: 10pt
 
@@
 conference:
-  name: PaPoC 2015
+  name: PaPoC'15
   location: 'April 21, 2015, Bordeaux, France'
   year: 2015
+  copyrightdata: 978-1-4503-3238-5/15/04
 
-doi: 0
+exclusivelicense: true
+doi: 2745947.2745951
 

### 2015-03-14 rmd/paper.Rmd
move things in appendix into the body

@@
 To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations.
-The transaction protocol employs standard two-phase commit, and two-phase locking; more details are in the appendix (\autoref{apx:protocol}).
+Our experiments were carried out with 4 shards on 4 local nodes, each with 8-core 2GHz Xeon E5335 processors and standard ethernet connecting them.
+
+## Transaction protocol
+
+The transaction protocol employs standard two-phase commit and two-phase locking with retries to ensure isolation, atomicity.
+We implement a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
+However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
+
+To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
+During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
+an "increment" speculatively returns the incremented value).
+When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
+This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
 
@@
 
-These experiments were run with 4 shards on 4 local nodes, each with 8-core 2GHz Xeon E5335 processors.
+## Microbenchmark: Set operations
 
@@
 
-**Microbenchmark: Set operations.** We first evaluate performance with a simple workload consisting of a raw mix of `Set` operations randomly distributed over 10,000 keys.
+We first evaluate performance with a simple workload consisting of a raw mix of `Set` operations randomly distributed over 10,000 keys.
 We use both a uniform random distribution as well as a skewed Zipfian distribution with a coefficient of 0.6.
@@
 
-```{r throughput, include=F, fig.height=2.1}
-d <- data(db("
-    select * from tapir where 
-    generator_time is not null and total_time is not null
-    and name like 'claret-v0.14%'
-  ",
-    factors=c('nshards', 'nclients'),
-    numeric=c('total_time', 'txn_count')
-))
-d.u <- subset(d, nshards == 4 & initusers == 4096 
-                & grepl('geom_repost|read_heavy', mix))
-ggplot(d.u, aes(x=nclients, y=throughput/1000, 
-        group=cc, fill=cc, color=cc, linetype=cc))+
-    stat_summary(fun.y=mean, geom="line", size=0.4)+
-    xlab('Concurrent clients')+ylab('Throughput (k trans. / sec)')+
-    expand_limits(y=0)+
-    facet_wrap(~workload)+
-    theme_mine+
-    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
-    cc_scales(title='Concurrency\ncontrol:')
-```
-\begin{figure}[t]
-\centering
-\includegraphics{figure/throughput-1.pdf}
-\caption{Throughput of social network workload (Retwis) with 4000 users.
-Leveraging commutativity prevents performance from falling over even
-when posts spread virally (repost-heavy).\label{tput}}
-\end{figure}
-
-**Case study: Retwis.** 
+## Case study: Retwis
 To understand performance on a typical web workload, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis [@redis].
 We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
-On top of Retwis's basic functionality, we also add a "repost" action that behaves like Twitter's "retweet".
-We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
-More about this workload is in the appendix (\autoref{apx:retwis}).
+On top of Retwis's basic functionality, we added a "repost" action that behaves like Twitter's "retweet".
 
-\autoref{tput} shows the results of this simulation.
-When most of the traffic is content consumption (reading timelines), both systems perform well enough.
-However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
-As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record [@ellenselfie], so too does our baseline fall over.
-But with commutativity, performance continues to scale even under this highly contentious load.
-
-# Appendix
-
-Some details did not fit into the core of the paper, but because they may help interpret our findings and give a sense of future directions, we include them here for anyone interested.
-
-## Transaction protocol {#apx:protocol}
-
-Our protocol for executing transactions is fairly straightforward: two-phase commit, and uses two-phase locking with retries to guarantee isolation.
-We employ a number of standard optimizations, such as delaying acquiring locks for operations that don't return a value to the *prepare* step so that locks are held for as short a time as possible.
-However, there is one step that is non-standard in order to support complex data types where rolling back state changes would be non-trivial.
-
-To support transactions with arbitrary data structure operations, each operation is split into two steps: *stage* and *apply*.
-During transaction execution, each operation's *stage* method attempts to acquire the necessary lock and may return a value *as if the operation has completed* (e.g.
-an "increment" speculatively returns the incremented value).
-When the transaction is prepared to commit, *apply* is called on each staged operation to actually mutate the underlying data structure.
-This allows operations to easily be un-staged if the transaction fails to acquire all the necessary locks, without requiring rollbacks.
-
-## Retwis workload {#apx:retwis}
 ```{r}
@@
 
-With our Retwis workload, rather than approximating the skew in real-world workloads with a Zipfian as many other systems do, we attempt to simulate the behavior of social networks with a realistic synthetic graph and a simple model of user behavior for posting and reposting.
+We simulate a realistic workload using a synthetic graph with power-law degree distribution and clients that randomly select between Retwis transactions including "add follower", "new post", and "repost", executing them as fast as they can.
+
+Rather than simply approximating the skew in real-world workloads with a Zipfian distribution as many other systems do, we simulate the behavior of social networks with a realistic synthetic graph and a simple model of user behavior for posting and reposting.
 
@@
 This generator is designed to result in graphs with the same power-law degree distribution found in natural graphs.
-\autoref{followers} shows the cumulative distribution function (CDF) of the number of followers per user for the synthetic graph of approximately 4000 users, with an average number of followers of 16 (scale 12 with edgefactor of 16 in Graph500's terms) which we use in our performance results back in \autoref{eval}.
+\autoref{followers} shows the cumulative distribution function (CDF) of the number of followers per user for the synthetic graph of approximately 4000 users, with an average number of followers of 16 (scale 12 with edgefactor of 16 in Graph500's terms).
 Most users should have relatively few followers; we see that roughly 50% have fewer than 100 followers, while a very small number of users have over 1000 followers.
@@
 Note that a small number of posts are reposted so much that they end up on over a quarter of users' timelines.
+
+```{r throughput, include=F, fig.height=2.1}
+d <- data(db("
+    select * from tapir where 
+    generator_time is not null and total_time is not null
+    and name like 'claret-v0.14%'
+  ",
+    factors=c('nshards', 'nclients'),
+    numeric=c('total_time', 'txn_count')
+))
+d.u <- subset(d, nshards == 4 & initusers == 4096 
+                & grepl('geom_repost|read_heavy', mix))
+ggplot(d.u, aes(x=nclients, y=throughput/1000, 
+        group=cc, fill=cc, color=cc, linetype=cc))+
+    stat_summary(fun.y=mean, geom="line", size=0.4)+
+    xlab('Concurrent clients')+ylab('Throughput (k trans. / sec)')+
+    expand_limits(y=0)+
+    facet_wrap(~workload)+
+    theme_mine+
+    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
+    cc_scales(title='Concurrency\ncontrol:')
+```
+\begin{figure}[t]
+\centering
+\includegraphics{figure/throughput-1.pdf}
+\caption{Throughput of social network workload (Retwis) with 4000 users.
+Leveraging commutativity prevents performance from falling over even
+when posts spread virally (repost-heavy).\label{tput}}
+\end{figure}
+
+\autoref{tput} shows the results of this simulation.
+When most of the traffic is content consumption (reading timelines), both systems perform well enough.
+However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post.
+As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record [@ellenselfie], so too does our baseline fall over.
+But with commutativity, performance continues to scale even under this highly contentious load.

### 2015-03-14 rmd/paper.Rmd
tweak figure 1

@@
 
-```{r stress, include=F, fig.height=2.4}
+```{r stress, include=F, fig.width=2.8, fig.height=3.1}
 d <- data(db("
@@
     theme_mine+
-    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
+    theme(legend.position=c(0.5,-0.22), plot.margin=unit(c(.5,.5,8,.5),'mm'), legend.direction='horizontal', legend.title.align=1)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 ))
-d.u <- subset(d, nshards == 4 & initusers == 4096 
+d.u <- subset(d, nshards == 4 & initusers == 4096 & nclients != 96 & nclients != 128
                 & grepl('geom_repost|read_heavy', mix))

### 2015-03-14 rmd/paper.Rmd
fix stupid biblio stuff

@@
 ## Case study: Retwis
-To understand performance on a typical web workload, we implemented a version of *Retwis*, a simplified Twitter clone designed originally for Redis [@redis].
-We use data structures such as sets to track each user's followers and posts, and keep a materialized up-to-date timeline for each user (represented as a sorted set).
+To understand performance on a typical web workload, we use *Retwis*, a simplified Twitter clone designed originally for Redis [@redis].
+Data structures such as sets are used track each user's followers and posts and keep a materialized up-to-date timeline for each user (represented as a sorted set).
 On top of Retwis's basic functionality, we added a "repost" action that behaves like Twitter's "retweet".
@@
 
-```{r followers, include=F, fig.width=2.5, fig.height=2.5}
+```{r followers, include=F, fig.width=2.5, fig.height=2.3}
 d.follow <- histogram.facets(subset(df,
@@
 
-```{r reposts, include=F, fig.width=2.5, fig.height=2.5}
+```{r reposts, include=F, fig.width=2.5, fig.height=2.3}
 d.repost <- histogram.facets(

### 2015-03-14 rmd/paper.Rmd
adjust figure 4 (move legend to bottom)

@@
 
-```{r throughput, include=F, fig.height=2.1}
+```{r throughput, include=F, fig.width=2.5, fig.height=2.8}
 d <- data(db("
@@
     theme_mine+
-    theme(legend.position='right', legend.direction='vertical', legend.title.align=0)+
+    theme(legend.position=c(0.5,-0.25), plot.margin=unit(c(.5,.5,8,.5),'mm'), legend.direction='horizontal', legend.title.align=1)+
     cc_scales(title='Concurrency\ncontrol:')

### 2015-03-24 data/common.R
reorganize repository to be 'claret-doc', remove now-outdated Authorea stuff

@@ -1,1 +1,159 @@
-../rmd/common.R
+suppressPackageStartupMessages(require(RMySQL))
+suppressPackageStartupMessages(require(sqldf))
+suppressPackageStartupMessages(require(ggplot2))
+suppressPackageStartupMessages(require(reshape))
+options(RMySQL.dbname="claret") # (rest comes from $HOME/.my.cnf)
+require(grid)
+library(jsonlite)
+library(scales)
+require(plyr)
+
+json.to.df <- function(jstr) {
+  d <- fromJSON(jstr)
+  return(data.frame(x=names(d),y=unlist(d)))
+}
+
+
+db <- function(query, factors=c(), numeric=c()) {
+  d <- sqldf(query)
+  d[factors] <- lapply(d[factors], factor)
+  d[numeric] <- lapply(d[numeric], as.numeric)
+  return(d)
+}
+
+num <- function(var) as.numeric(as.character(var))
+
+
+df.histogram <- function(json, version="none") {
+  d <- fromJSON(json)
+  return(data.frame(x=num(names(d)), y=num(unlist(d)), version=version))
+}
+
+
+save <- function(g, name=FILE_BASE, file=sprintf("%s/%s.pdf",FILE_DIR,name), w=3.3, h=3.1) {
+  ggsave(plot=g, filename=file, width=w, height=h)
+  print(sprintf("saved: %s", file))
+}
+
+prettify <- function(str) gsub('_',' ',gsub('([a-z])([a-z]+)',"\\U\\1\\E\\2",str,perl=TRUE))
+
+regex_match <- function(reg,str) length(grep(reg,str)) > 0
+
+label_pretty <- function(variable, value) {
+  vname <- if (regex_match('variable|value',variable)) '' else sprintf('%s:', variable)
+  lapply(paste(vname, prettify(as.character(value))), paste, collapse="\n")
+}
+
+geom_mean <- function(geom) stat_summary(fun.y='mean', geom=geom, labeller=label_pretty)
+
+geom_meanbar <- function(labeller=label_pretty) {
+  return(list(
+    stat_summary(fun.y=mean, geom='bar', position='dodge'),
+    stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, position='dodge')
+  ))
+}
+
+c.blue   <- "#0072B2"
+c.yellow <- "#E69F00"
+c.green  <- "#009E73"
+c.red    <- "#D55E00"
+c.pink   <- "#CC79A7"
+c.gray   <- "#999999"
+
[…]

### 2015-03-24 papoc/common.R

@@ -1,0 +1,1 @@
+../data/common.R

### 2015-03-24 rmd/common.R

@@ -1,159 +1,0 @@
-suppressPackageStartupMessages(require(RMySQL))
-suppressPackageStartupMessages(require(sqldf))
-suppressPackageStartupMessages(require(ggplot2))
-suppressPackageStartupMessages(require(reshape))
-options(RMySQL.dbname="claret") # (rest comes from $HOME/.my.cnf)
-require(grid)
-library(jsonlite)
-library(scales)
-require(plyr)
-
-json.to.df <- function(jstr) {
-  d <- fromJSON(jstr)
-  return(data.frame(x=names(d),y=unlist(d)))
-}
-
-
-db <- function(query, factors=c(), numeric=c()) {
-  d <- sqldf(query)
-  d[factors] <- lapply(d[factors], factor)
-  d[numeric] <- lapply(d[numeric], as.numeric)
-  return(d)
-}
-
-num <- function(var) as.numeric(as.character(var))
-
-
-df.histogram <- function(json, version="none") {
-  d <- fromJSON(json)
-  return(data.frame(x=num(names(d)), y=num(unlist(d)), version=version))
-}
-
-
-save <- function(g, name=FILE_BASE, file=sprintf("%s/%s.pdf",FILE_DIR,name), w=3.3, h=3.1) {
-  ggsave(plot=g, filename=file, width=w, height=h)
-  print(sprintf("saved: %s", file))
-}
-
-prettify <- function(str) gsub('_',' ',gsub('([a-z])([a-z]+)',"\\U\\1\\E\\2",str,perl=TRUE))
-
-regex_match <- function(reg,str) length(grep(reg,str)) > 0
-
-label_pretty <- function(variable, value) {
-  vname <- if (regex_match('variable|value',variable)) '' else sprintf('%s:', variable)
-  lapply(paste(vname, prettify(as.character(value))), paste, collapse="\n")
-}
-
-geom_mean <- function(geom) stat_summary(fun.y='mean', geom=geom, labeller=label_pretty)
-
-geom_meanbar <- function(labeller=label_pretty) {
-  return(list(
-    stat_summary(fun.y=mean, geom='bar', position='dodge'),
-    stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, position='dodge')
-  ))
-}
-
-c.blue   <- "#0072B2"
-c.yellow <- "#E69F00"
-c.green  <- "#009E73"
-c.red    <- "#D55E00"
-c.pink   <- "#CC79A7"
-c.gray   <- "#999999"
-
-my_palette <- c(
[…]

### 2015-03-24 data/common.R
copy notes from other Authorea project

@@ -3,10 +3,9 @@ suppressPackageStartupMessages(require(sqldf))
 suppressPackageStartupMessages(require(ggplot2))
 suppressPackageStartupMessages(require(reshape))
 options(RMySQL.dbname="claret") # (rest comes from $HOME/.my.cnf)
-require(grid)
-library(jsonlite)
-library(scales)
-require(plyr)
+
+suppressPackageStartupMessages(library(jsonlite))
+suppressPackageStartupMessages(library(scales))
 
 json.to.df <- function(jstr) {
   d <- fromJSON(jstr)
@@ -23,6 +22,7 @@ db <- function(query, factors=c(), numeric=c()) {
 
 num <- function(var) as.numeric(as.character(var))
 
+capply <- function(col, func) unlist(lapply(col, func))
 
 df.histogram <- function(json, version="none") {
   d <- fromJSON(json)
@@ -66,6 +66,9 @@ my_palette <- c(
   
   'reader/writer'=c.yellow,
   'commutative'=c.blue,
+
+  'approx'=c.green,
+  'precise'=c.blue,
   
   'follow'=c.blue,
   'newuser'=c.yellow,
@@ -96,6 +99,20 @@ cc_scales <- function(field=cc, title="Concurrency control:") list(
   scale_linetype_manual(name=title, values=c('commutative'=1,'reader/writer'=2))
 )
 
+my_theme <- function() theme(
+  panel.background = element_rect(fill="white"),
+  panel.border = element_rect(fill=NA, color="grey50"),
+  panel.grid.major = element_line(color="grey80", size=0.2),
+  panel.grid.minor = element_line(color="grey90", size=0.2),
+  strip.background = element_rect(fill="grey90", color="grey50"),
+  strip.background = element_rect(fill="grey80", color="grey50"),
+  axis.ticks = element_line(colour="black"),
+  panel.grid = element_line(colour="black"),
+  axis.text.y = element_text(colour="black"),
+  axis.text.x = element_text(colour="black"),
+  text = element_text(size=10, family="Helvetica")
+)
+
 
 theme_mine <- list(
   scale_fill_manual(values=my_palette),
@@ -103,25 +120,17 @@ theme_mine <- list(
   scale_color_manual(values=my_palette),
   # To use for fills, add
   # basic black and white theme
-  theme(
-    panel.background = element_rect(fill="white"),
[…]

### 2015-03-24 data/histograms.R

@@ -1,0 +1,75 @@
+#!/usr/bin/env Rscript
+source('common.R')
+
+df <- subset(db("select * from tapir where stat_following_counts is not null and name like '%v0.14%'"),
+  nclients == 32
+  & initusers == 4096
+)
+df$grp <- with(df, sprintf("%s\n%s\nmix:%s/%s,\n%s", name, ccmode, mix, alpha, gen))
+
+
+histogram.facets <- function(df, measure, grp) {
+  d <- data.frame(x=c(),y=c(),version=c())
+  for (i in 1:nrow(df)) {
+    d <- rbind(d, df.histogram(df[i,measure], df[i,grp]))
+  }
+  return(d)
+}
+
+d.repost <- histogram.facets(subset(df, initusers == 4096 & mix == 'geom_repost'), 'stat_repost_counts', 'grp')
+save(
+  ggplot(d.repost, aes(x=x, weight=y))+
+    # stat_summary(fun.y=sum, geom='bar', fill=c.blue)+
+    # geom_bar(stat="identity", fill=c.blue)+
+    # geom_histogram(binwidth=.1, fill=c.blue)+
+    stat_ecdf(color=c.blue)+
+    xlab('# reposts')+ylab('count')+
+    scale_x_log10(breaks=c(1,10,100,1000))+scale_y_log10(breaks=c(0.1,0.2,0.4,0.6,0.8,1.0))+
+    # scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+    #               labels = trans_format("log10", function(x) round(10^x, digits=2)))+
+    # scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+    #               labels = trans_format("log10", function(x) round(10^x, digits=2)))+
+    
+    xlab('# reposts (log scale)')+ylab('CDF (log scale)')+
+    # facet_wrap(~version, scale="free")+
+    theme_mine
+, name='hist_reposts', w=4, h=4)
+
+
+d.post <- histogram.facets(df, 'stat_post_counts', 'grp')
+save(
+  ggplot(d.post, aes(x=x, weight=y))+
+    # stat_summary(fun.y=sum, geom='bar', fill=c.blue)+
+    # geom_bar(stat="identity", fill=c.blue)+
+    geom_histogram(binwidth=1, fill=c.blue)+
+    xlab('# posts / user')+ylab('count')+
+    # scale_y_log10()+
[…]

### 2015-03-24 data/ldbc.R

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+# setwd('/Users/bholt/hub/claret/data')
+source('common.R')
+
+# DATA.MODE <- 'local'
+d <- data.ldbc(where="name like '%-multinode'")
+# d$ntotal <- factor(d$ntotal)
+
+save(
+  ggplot(subset(d,
+    # ccmode == 'simple'
+    ntotal == 50000
+    # & name == 'Query2'
+  ), aes(
+    x = snb_time_ratio,
+    y = total_time / 1e6,
+    group = cc,
+    fill = cc,
+    color = cc,
+    label = ntotal,
+  ))+
+  geom_point()+
+  # geom_text(size=1.7)+
+  # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
+  stat_smooth()+
+  expand_limits(y=0)+
+  facet_wrap(~name)+
+  scale_x_log10(breaks=trans_breaks("log10", function(x) 10^x),
+                labels=trans_format("log10", math_format(10^.x)))+
+  cc_scales()+
+  my_theme()
+  # scale_fill_manual(values=my_palette, name='Variant')+
+  # scale_color_manual(values=my_palette, name='Variant')
+, name='explore_ldbc', w=10, h=8)
+
+save(
+  ggplot(subset(d,
+    # ccmode == 'simple'
+    ntotal == 50000
+    # & name == 'AddPost'
+  ), aes(
+    x = snb_time_ratio,
+    y = time_mean,
+    group = cc,
+    fill = cc,
+    color = cc,
+    label = ntotal,
+  ))+
+  geom_point()+
+  # geom_text(size=1.7)+
+  # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
+  stat_smooth()+
+  expand_limits(y=0)+
+  facet_wrap(~name)+
+  scale_x_log10(breaks=trans_breaks("log10", function(x) 10^x),
+                labels=trans_format("log10", math_format(10^.x)))+
+  cc_scales()+
+  my_theme()
+, name='explore_latency', w=10, h=8)
+
+save(
+  ggplot(subset(d,
+    # ccmode == 'simple'
+    ntotal == 50000
+    # & name == 'Query2'
+  ), aes(
+    x = snb_time_ratio,
+    y = server_cc_check_success,
+    group = cc,
+    fill = cc,
+    color = cc,
+    label = ntotal,
+  ))+
+  geom_point()+
[…]

### 2015-03-24 data/plot.R

@@ -1,0 +1,433 @@
+#!/usr/bin/env Rscript
+source('common.R')
+
+my_smooth <- function() stat_smooth(method=loess) #, span=0.3)
+
+capply <- function(col, func) unlist(lapply(col, func))
+
+data <- function(d) {
+  d$failure_rate <- d$txn_failed / (d$txn_count + d$txn_failed)
+  d$throughput <- d$txn_count * num(d$nclients) / d$total_time
+  # d$throughput <- d$ntxns * num(d$nclients) / d$total_time
+  d$avg_latency_ms <- d$txn_time / d$txn_count * 1000
+  
+  d$prepare_total <- d$prepare_retries + d$txn_count
+  d$prepare_retry_rate <- d$prepare_retries / d$prepare_total
+  
+  # d$cc <- revalue(d$ccmode, c(
+  #   'bottom'='base (none)',
+  #   'rw'='reader/writer',
+  #   'simple'='commutative'
+  # ))
+  
+  d$ccf <- factor(d$ccmode, levels=c('simple','rw','bottom'))
+  
+  d$cc <- factor(revalue(d$ccmode, c(
+    # 'bottom'='base',
+    'rw'='reader/writer',
+    'simple'='commutative'
+  )), levels=c('commutative','reader/writer','base'))
+  d$`Concurrency Control` <- d$cc
+  
+  d$Graph <- capply(d$gen, function(s) gsub('kronecker:.+','kronecker',s))
+  
+  
+  d$workload <- factor(revalue(d$mix, c(
+    'geom_repost'='repost-heavy',
+    'read_heavy'='read-heavy',
+    'update_heavy'='mixed'
+  )), levels=c('repost-heavy','read-heavy','mixed'))
+  
+  d$zmix <- sprintf('%s/%s', d$mix, d$alpha)
+  
+  d$facet <- sprintf('shards: %d\n%s\n%d users\n%s', num(d$nshards), d$zmix, d$initusers, d$Graph)
+
+  d$gen_label <- sprintf('%d users\n%s', d$initusers, d$Graph)
+  
+  return(d)
+}
+#
+# d.all <- data(db("
+#   select * from tapir where
+#   generator_time is not null and total_time is not null
+#   and name like 'claret-v%'
+#   and nshards = 4
+# ",
+#   factors=c('nshards', 'nclients'),
+#   numeric=c('total_time', 'txn_count')
+# ))
+
+d <- data(db("
+  select * from tapir where 
[…]

### 2015-03-25 papoc/paper.Rmd
reorganize repository to be 'claret-doc', remove now-outdated Authorea stuff
fix papoc data functions and tweak figures

--- 2015-03-14 rmd/paper.Rmd
@@
 layout: sigplanconf
-bibliography: biblio.bib
+bibliography: refs.bib
 output:
@@
 ```{r stress, include=F, fig.width=2.8, fig.height=3.1}
-d <- data(db("
-    select * from tapir where 
-    total_time is not null
-    and name like 'stress-v0.14%'
-    ", factors=c('nshards', 'nclients'), numeric=c('total_time', 'txn_count')))
+d <- data.papoc(where="name like 'stress-v0.14%'")
 d$opmix <- factor(revalue(d$mix, c(
@@
     theme_mine+
-    theme(legend.position=c(0.5,-0.22), plot.margin=unit(c(.5,.5,8,.5),'mm'), legend.direction='horizontal', legend.title.align=1)+
+    theme(legend.position=c(0.5,-0.25), plot.margin=unit(c(.5,.5,10,.5),'mm'), legend.direction='horizontal', legend.title.align=1)+
     cc_scales(title='Concurrency\ncontrol:')
@@
 
-```{r throughput, include=F, fig.width=2.5, fig.height=2.8}
-d <- data(db("
-    select * from tapir where 
-    generator_time is not null and total_time is not null
-    and name like 'claret-v0.14%'
-  ",
-    factors=c('nshards', 'nclients'),
-    numeric=c('total_time', 'txn_count')
-))
+```{r throughput, include=F, fig.width=2.8, fig.height=2.8}
+d <- data.papoc(where="name like 'claret-v0.14%'")
 d.u <- subset(d, nshards == 4 & initusers == 4096 & nclients != 96 & nclients != 128

### 2015-03-25 data/common.R

@@ -6,6 +6,7 @@ options(RMySQL.dbname="claret") # (rest comes from $HOME/.my.cnf)
 
 suppressPackageStartupMessages(library(jsonlite))
 suppressPackageStartupMessages(library(scales))
+suppressPackageStartupMessages(require(grid))
 
 json.to.df <- function(jstr) {
   d <- fromJSON(jstr)
@@ -110,7 +111,7 @@ my_theme <- function() theme(
   panel.grid = element_line(colour="black"),
   axis.text.y = element_text(colour="black"),
   axis.text.x = element_text(colour="black"),
-  text = element_text(size=10, family="Helvetica")
+  text = element_text(size=9, family="Helvetica")
 )
 
 
@@ -123,8 +124,7 @@ theme_mine <- list(
   my_theme()
 )
 
-
-claret_data <- function(where) {
+data.papoc <- function(where) {
   d <- db(
     sprintf("select * from tapir where total_time is not null and %s", where),
     factors=c('nshards', 'nclients'),

### 2015-03-25 data/common.R
seems like we need to require 'plyr'

@@ -7,6 +7,7 @@ options(RMySQL.dbname="claret") # (rest comes from $HOME/.my.cnf)
 suppressPackageStartupMessages(library(jsonlite))
 suppressPackageStartupMessages(library(scales))
 suppressPackageStartupMessages(require(grid))
+suppressPackageStartupMessages(require(plyr))
 
 json.to.df <- function(jstr) {
   d <- fromJSON(jstr)

### 2015-03-30 data/common.R
add retwis lat v throughput plots

@@ -125,6 +125,41 @@ theme_mine <- list(
   my_theme()
 )
 
+data.retwis <- function(where="client = 'dsretwis'") {
+  d <- db(
+    sprintf("select * from retwis where total_time is not null and %s", where),
+    factors=c('nshards', 'nclients'),
+    numeric=c('total_time', 'txn_count')
+  )
+  
+  d$throughput <- d$txn_count * num(d$nclients) / d$total_time
+  # d$throughput <- d$ntxns * num(d$nclients) / d$total_time
+  d$avg_latency_ms <- d$txn_time / d$txn_count * 1000
+
+  d$prepare_total <- d$prepare_retries + d$txn_count
+  d$prepare_retry_rate <- d$prepare_retries / d$prepare_total
+  
+  d$cc <- factor(revalue(d$ccmode, c(
+    'rw'='reader/writer',
+    'simple'='commutative'
+  )), levels=c('commutative','reader/writer'))
+  d$`Concurrency Control` <- d$cc
+  
+  d$graph <- mapply(function(g,d){ if(g == 'none') gsub('^.*/kronecker/([0-9]+)','kronecker:\\1',d) else g }, d$gen, d$loaddir)
+  
+  d$workload <- factor(revalue(d$mix, c(
+    'geom_repost'='repost-heavy',
+    'read_heavy'='read-heavy',
+    'update_heavy'='mixed'
+  )), levels=c('repost-heavy','read-heavy','mixed'))
+  
+  d$zmix <- sprintf('%s/%s', d$mix, d$alpha)
+
+  d$facet <- sprintf('%s\n%s', d$zmix, d$graph)
+  
+  return(d)
+}
+
 data.papoc <- function(where) {
   d <- db(
     sprintf("select * from tapir where total_time is not null and %s", where),

### 2015-03-30 data/retwis.R

@@ -1,0 +1,90 @@
+#!/usr/bin/env Rscript
+source('common.R')
+
+d <- data.retwis(where="nshards = 4 and rate != 0 and loaddir like '%12%'")
+
+sub <- function(d=d) subset(d, mix == 'geom_repost')
+
+# save(
+#   ggplot(melt(
+#     sub(d),
+#     measure=c('throughput','avg_latency_ms')
+#   ), aes(
+#     x = rate,
+#     y = value,
+#     group = cc,
+#     fill = cc,
+#     color = cc,
+#     label = rate,
+#   ))+
+#   geom_point()+
+#   # geom_text(size=1.7)+
+#   # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
+#   # stat_smooth()+
+#   expand_limits(y=0)+
+#   facet_wrap(~variable, scales="free_y")+
+#   # scale_x_log10(breaks=trans_breaks("log10", function(x) 10^x),
+#   #               labels=trans_format("log10", math_format(10^.x)))+
+#   cc_scales()+
+#   my_theme()
+# , name='plot/retwis', w=10, h=8)
+
+d.mean <- ddply(d, .(workload,cc,rate), summarize, x=mean(throughput), y=mean(avg_latency_ms))
+
+save(
+  ggplot(d, aes(
+    x = throughput,
+    y = avg_latency_ms,
+    group = cc,
+    fill = cc,
+    color = cc,
+  ))+
+  xlab('Throughput')+ylab('Average latency (ms)')+
+  geom_point()+
+  geom_path(data=d.mean, aes(x=x,y=y))+
+  # geom_path(stat=summary, stat_params=list(fun.x=mean, fun.y=mean), aes(x=x,y=y))+
+  expand_limits(y=0)+
+  cc_scales()+
+  facet_wrap(~workload)+
+  my_theme()
+, name='plot/retwis_tput_v_lat', w=8, h=5)
+
+
+d$retwis_newuser_latency  <- d$retwis_newuser_time  / d$retwis_newuser_count
+d$retwis_post_latency     <- d$retwis_post_time     / d$retwis_post_count
+d$retwis_repost_latency   <- d$retwis_repost_time   / d$retwis_repost_count
+d$retwis_timeline_latency <- d$retwis_timeline_time / d$retwis_timeline_count
+d$retwis_follow_latency   <- d$retwis_follow_time   / d$retwis_follow_count
+
+d.m <- melt(sub(d),
[…]

### 2015-04-01 data/ldbc.R
working with ldbc

@@ -1,10 +1,9 @@
 #!/usr/bin/env Rscript
-# setwd('/Users/bholt/hub/claret/data')
 source('common.R')
 
 # DATA.MODE <- 'local'
-d <- data.ldbc(where="name like '%-multinode'")
-# d$ntotal <- factor(d$ntotal)
+# d <- data.ldbc(where="name like '%v0.16-multinode%' and snb_threads = 24")
+d <- data.ldbc()
 
 save(
   ggplot(subset(d,
@@ -12,31 +11,28 @@ save(
     ntotal == 50000
     # & name == 'Query2'
   ), aes(
-    x = snb_time_ratio,
-    y = total_time / 1e6,
+    x = 1/snb_time_ratio,
+    y = throughput,
     group = cc,
     fill = cc,
     color = cc,
     label = ntotal,
   ))+
   geom_point()+
-  # geom_text(size=1.7)+
-  # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
-  stat_smooth()+
+  # stat_smooth()+
   expand_limits(y=0)+
-  facet_wrap(~name)+
+  facet_wrap(~machines)+
   scale_x_log10(breaks=trans_breaks("log10", function(x) 10^x),
                 labels=trans_format("log10", math_format(10^.x)))+
   cc_scales()+
   my_theme()
-  # scale_fill_manual(values=my_palette, name='Variant')+
-  # scale_color_manual(values=my_palette, name='Variant')
 , name='explore_ldbc', w=10, h=8)
 
 save(
   ggplot(subset(d,
     # ccmode == 'simple'
     ntotal == 50000
+    & grepl('zork',machines)
     # & name == 'AddPost'
   ), aes(
     x = snb_time_ratio,

### 2015-04-01 data/common.R
separate plots for external client (zork) vs. clients-on-same-nodes

@@ -24,6 +24,8 @@ db <- function(query, factors=c(), numeric=c()) {
 
 num <- function(var) as.numeric(as.character(var))
 
+x <- function(...) { return(paste(..., sep='#')) }
+
 capply <- function(col, func) unlist(lapply(col, func))
 
 df.histogram <- function(json, version="none") {
@@ -114,7 +116,6 @@ my_theme <- function() theme(
   axis.text.x = element_text(colour="black"),
   text = element_text(size=9, family="Helvetica")
 )
-
 
 theme_mine <- list(
   scale_fill_manual(values=my_palette),
@@ -231,7 +232,7 @@ data.ldbc <- function(where = "ldbc_config is not null") {
     colnames(mr) <- sprintf("time_%s", colnames(mr))
     
     data.frame(
-      throughput=as.numeric(r$ntotal)/as.numeric(o$total_duration),
+      throughput=as.numeric(r$ntotal)/as.numeric(o$total_duration)*1e6,
       total_time=o$total_duration,
       name=o$all_metrics$name,
       count=o$all_metrics$count,

### 2015-04-01 data/retwis.R

@@ -1,7 +1,7 @@
 #!/usr/bin/env Rscript
 source('common.R')
 
-d <- data.retwis(where="nshards = 4 and rate != 0 and loaddir like '%12%'")
+d <- data.retwis(where="nshards = 4 and rate != 0 and loaddir like '%12%' and rate < 1000")
 
 sub <- function(d=d) subset(d, mix == 'geom_repost')
 
@@ -29,62 +29,73 @@ sub <- function(d=d) subset(d, mix == 'geom_repost')
 #   my_theme()
 # , name='plot/retwis', w=10, h=8)
 
-d.mean <- ddply(d, .(workload,cc,rate), summarize, x=mean(throughput), y=mean(avg_latency_ms))
+d$throughput <- d$retwis_txn_count * num(d$nclients) / d$total_time;
 
-save(
-  ggplot(d, aes(
-    x = throughput,
-    y = avg_latency_ms,
-    group = cc,
-    fill = cc,
-    color = cc,
-  ))+
-  xlab('Throughput')+ylab('Average latency (ms)')+
-  geom_point()+
-  geom_path(data=d.mean, aes(x=x,y=y))+
-  # geom_path(stat=summary, stat_params=list(fun.x=mean, fun.y=mean), aes(x=x,y=y))+
-  expand_limits(y=0)+
-  cc_scales()+
-  facet_wrap(~workload)+
-  my_theme()
-, name='plot/retwis_tput_v_lat', w=8, h=5)
+d$client_machines <- factor(d$machines)
 
+plot.path <- function(data) {
+  d.mean <- ddply(data, .(workload,cc,rate), summarize, x=mean(throughput), y=mean(avg_latency_ms))
 
-d$retwis_newuser_latency  <- d$retwis_newuser_time  / d$retwis_newuser_count
-d$retwis_post_latency     <- d$retwis_post_time     / d$retwis_post_count
-d$retwis_repost_latency   <- d$retwis_repost_time   / d$retwis_repost_count
-d$retwis_timeline_latency <- d$retwis_timeline_time / d$retwis_timeline_count
-d$retwis_follow_latency   <- d$retwis_follow_time   / d$retwis_follow_count
-
-d.m <- melt(sub(d),
-  measure=c(
-    'retwis_newuser_latency',
-    'retwis_post_latency',
-    'retwis_repost_latency',
-    'retwis_timeline_latency',
-    'retwis_follow_latency'
+  return(
+    ggplot(data, aes(
[…]

### 2015-04-01 data/ldbc.R
put ldbc plots in plot/

@@ -26,7 +26,7 @@ save(
                 labels=trans_format("log10", math_format(10^.x)))+
   cc_scales()+
   my_theme()
-, name='explore_ldbc', w=10, h=8)
+, name='plot/explore_ldbc', w=10, h=8)
 
 save(
   ggplot(subset(d,
@@ -52,7 +52,7 @@ save(
                 labels=trans_format("log10", math_format(10^.x)))+
   cc_scales()+
   my_theme()
-, name='explore_latency', w=10, h=8)
+, name='plot/explore_latency', w=10, h=8)
 
 save(
   ggplot(subset(d,
@@ -79,7 +79,7 @@ save(
   my_theme()
   # scale_fill_manual(values=my_palette, name='Variant')+
   # scale_color_manual(values=my_palette, name='Variant')
-, name='explore_retries', w=10, h=8)
+, name='plot/explore_retries', w=10, h=8)
 
 save(
   ggplot(subset(d, ntotal==50000), aes(
@@ -93,4 +93,4 @@ save(
   theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0))+
   expand_limits(y=0)+
   my_theme()
-, name='explore_counts', w=8, h=6)
+, name='plot/explore_counts', w=8, h=6)

### 2015-04-02 data/ldbc.R
show different machine configurations in facets

@@ -3,7 +3,7 @@ source('common.R')
 
 # DATA.MODE <- 'local'
 # d <- data.ldbc(where="name like '%v0.16-multinode%' and snb_threads = 24")
-d <- data.ldbc()
+d <- subset(data.ldbc(), is.na(machines) | grepl('zork.cs',machines))
 
 save(
   ggplot(subset(d,
@@ -32,17 +32,18 @@ save(
   ggplot(subset(d,
     # ccmode == 'simple'
     ntotal == 50000
-    & grepl('zork',machines)
+    & is.na(machines)
+    #& grepl('zork',machines)
     # & name == 'AddPost'
   ), aes(
-    x = snb_time_ratio,
-    y = time_mean,
+    x = 1/snb_time_ratio,
+    y = time_mean / 1000,
     group = cc,
     fill = cc,
     color = cc,
     label = ntotal,
   ))+
-  geom_point()+
+  geom_point()+ylab('avg latency (ms)')+
   # geom_text(size=1.7)+
   # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
   stat_smooth()+
@@ -58,10 +59,10 @@ save(
   ggplot(subset(d,
     # ccmode == 'simple'
     ntotal == 50000
-    # & name == 'Query2'
+    & name == 'Query2'
   ), aes(
     x = snb_time_ratio,
-    y = server_cc_check_success,
+    y = server_cc_check_failed / server_cc_check_count,
     group = cc,
     fill = cc,
     color = cc,
@@ -72,7 +73,7 @@ save(
   # stat_summary(aes(label=ntotal), fun.y=mean, geom="text")+
   stat_smooth()+
   # expand_limits(y=0)+
-  facet_wrap(~name)+
+  facet_wrap(~machines)+
   scale_x_log10(breaks=trans_breaks("log10", function(x) 10^x),
                 labels=trans_format("log10", math_format(10^.x)))+
   cc_scales()+

### 2015-04-02 data/retwis.R

@@ -1,9 +1,9 @@
 #!/usr/bin/env Rscript
 source('common.R')
 
-d <- data.retwis(where="nshards = 4 and rate != 0 and loaddir like '%12%' and rate < 1000")
+d <- data.retwis(where="nshards = 4 and rate != 0 and loaddir like '%12%' and rate <= 1000")
 
-sub <- function(d=d) subset(d, mix == 'geom_repost')
+sub <- function(d=d) subset(d, mix == 'geom_repost' & is.na(machines))
 
 # save(
 #   ggplot(melt(
@@ -33,11 +33,14 @@ d$throughput <- d$retwis_txn_count * num(d$nclients) / d$total_time;
 
 d$client_machines <- factor(d$machines)
 
-plot.path <- function(data) {
-  d.mean <- ddply(data, .(workload,cc,rate), summarize, x=mean(throughput), y=mean(avg_latency_ms))
+plot.path <- function(d) {
+  
+  d$facet <- sprintf('%s\n%s', d$machines, d$workload)
+  
+  d.mean <- ddply(d, .(facet,cc,rate), summarize, x=mean(throughput), y=mean(avg_latency_ms))
 
   return(
-    ggplot(data, aes(
+    ggplot(d, aes(
       x = throughput,
       y = avg_latency_ms,
       group = cc,
@@ -49,13 +52,15 @@ plot.path <- function(data) {
     geom_path(data=d.mean, aes(x=x,y=y))+
     expand_limits(y=0)+
     cc_scales()+
-    facet_wrap(~workload)+
+    facet_wrap(~facet, scales="free_x", ncol=3)+
     my_theme()
   )
 }
 
-save(plot.path(subset(d, !grepl('zork',machines))), name='plot/retwis_tput_v_lat', w=8, h=5)
-save(plot.path(subset(d, grepl('zork',machines))), name='plot/retwis_tput_v_lat_zork', w=8, h=5)
+# save(plot.path(subset(d, !grepl('zork',machines))), name='plot/retwis_tput_v_lat', w=8, h=5)
+save(plot.path(
+  subset(d, is.na(machines) | grepl('^candy$|,',machines) )
+), name='plot/retwis_tput_v_lat_zork', w=8, h=10)
 
 
 plot.path.breakdown <- function(d) {
@@ -64,8 +69,8 @@ plot.path.breakdown <- function(d) {
[…]

### 2015-09-22 full/plots/rawmix-tput-vs-lat.r
make rawmix-tput-vs-lat into a bar plot of peak throughput

@@ -24,39 +24,66 @@ d$x <- d$nthreads * num(d$nclients)
 g.cc_ph <- guide_legend(nrow = 6)
 
 
-save(
-  ggplot(d, aes(
-    x = throughput,
-    y = avg_latency_ms,
-    group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph
-  ))+
-  xlab('Throughput (txn/s)')+ylab('Mean latency (ms)')+
-  # geom_point()+
-  # geom_text(aes(label=label), size=1.7)+
-  scale_x_continuous(labels=si.labels())+
-  # geom_point()+
-  geom_mean_path(d, throughput, avg_latency_ms, .(x,cc,phasing,cc_ph))+
-  expand_limits(y=0)+
-  coord_cartesian(ylim=c(0,20))+
-  # cc_scales(title='Mode:', guide = g.cc)+
-  # color_scales('', my_palette)+
-  # phasing.linetype(title='Phasing:', guide = g.cc_ph)+
-  cc_ph_scales()+
-  my_theme() #+legend.bottom()
-, w=5, h=3)
+# save(
+#   ggplot(d, aes(
+#     x = throughput,
+#     y = avg_latency_ms,
+#     group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph
+#   ))+
+#   xlab('Throughput (txn/s)')+ylab('Mean latency (ms)')+
+#   # geom_point()+
+#   # geom_text(aes(label=label), size=1.7)+
+#   scale_x_continuous(labels=si.labels())+
+#   # geom_point()+
+#   geom_mean_path(d, throughput, avg_latency_ms, .(x,cc,phasing,cc_ph))+
+#   expand_limits(y=0)+
+#   coord_cartesian(ylim=c(0,20))+
+#   # cc_scales(title='Mode:', guide = g.cc)+
+#   # color_scales('', my_palette)+
+#   # phasing.linetype(title='Phasing:', guide = g.cc_ph)+
+#   cc_ph_scales()+
+#   my_theme() #+legend.bottom()
+# , w=5, h=3)
 
 
 # save(
-#   ggplot(d, aes(
+#   ggplot(subset(d, cc_ph != COMB), aes(
 #     x = x,
-#     y = txns_behind_schedule,
-#     group = x(cc,phasing),
-#     fill = cc, color = cc,
-#     linetype = phasing,
+#     y = throughput,
[…]

### 2015-09-22 full/plots/rawmix-tput-vs-lat.r
rawmix: add separate tput & latency plots

@@ -46,21 +46,39 @@ g.cc_ph <- guide_legend(nrow = 6)
 # , w=5, h=3)
 
 
-# save(
-#   ggplot(subset(d, cc_ph != COMB), aes(
-#     x = x,
-#     y = throughput,
-#     group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph,
-#   ))+
-#   xlab('total clients')+ylab('throughput (txn/s)')+
-#   stat_summary(geom='line', fun.y=mean)+
-#   # stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, aes(color='black'))+
-#   scale_x_continuous(trans=log2_trans(), breaks=c(16,32,64,128,256,384))+
-#   scale_y_continuous(labels=si.labels())+
-#   expand_limits(y=0)+
-#   cc_ph_scales()+
-#   my_theme()
-# , w=5, h=3.5)
+save(
+  ggplot(d, aes(
+    x = x,
+    y = throughput,
+    group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph,
+  ))+
+  xlab('total clients')+ylab('throughput (txn/s)')+
+  stat_summary(geom='line', fun.y=mean)+
+  # stat_summary(geom='smooth', fun.data=mean_cl_normal)+
+  # stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, aes(color='black'))+
+  scale_x_continuous(trans=log2_trans(), breaks=c(16,32,64,128,256,384))+
+  scale_y_continuous(labels=si.labels())+
+  expand_limits(y=0)+
+  cc_ph_scales()+
+  my_theme()
+, 'rawmix-tput', w=5, h=3.5)
+
+save(
+  ggplot(d, aes(
+    x = x,
+    y = avg_latency_ms,
+    group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph,
+  ))+
+  xlab('total clients')+ylab('throughput (txn/s)')+
+  stat_summary(geom='line', fun.y=mean)+
+  # stat_summary(geom='smooth', fun.data=mean_cl_normal)+
+  # stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, aes(color='black'))+
+  scale_x_continuous(breaks=c(16,32,64,128,256,384))+
+  scale_y_continuous(labels=si.labels())+
[…]

### 2015-09-22 data/common.r
make tput-vs-lat plots into clients-vs-tput

@@ -355,8 +355,8 @@ data.retwis <- function(select="*", where="client = 'dsretwis'") {
   
   
   lvl <- c(
-    'read-heavy (~14% post/repost)',
-    'post-heavy (~45% post/repost)',
+    'read-heavy: ~14% (re)post',
+    'post-heavy: ~45% (re)post',
     'mixed'
   )
   

### 2015-09-22 full/plots/rawmix-tput-vs-lat.r

@@ -52,10 +52,9 @@ save(
     y = throughput,
     group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph,
   ))+
-  xlab('total clients')+ylab('throughput (txn/s)')+
+  xlab('Clients')+ylab('Throughput (txn/s)')+
   stat_summary(geom='line', fun.y=mean)+
-  # stat_summary(geom='smooth', fun.data=mean_cl_normal)+
-  # stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, aes(color='black'))+
+  stat_summary(geom='point', fun.y=mean)+  
   scale_x_continuous(trans=log2_trans(), breaks=c(16,32,64,128,256,384))+
   scale_y_continuous(labels=si.labels())+
   expand_limits(y=0)+
@@ -69,8 +68,9 @@ save(
     y = avg_latency_ms,
     group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph,
   ))+
-  xlab('total clients')+ylab('throughput (txn/s)')+
+  xlab('Clients')+ylab('Mean transaction latency (ms)')+
   stat_summary(geom='line', fun.y=mean)+
+  stat_summary(geom='point', fun.y=mean)+  
   # stat_summary(geom='smooth', fun.data=mean_cl_normal)+
   # stat_summary(fun.data=mean_cl_normal, geom='errorbar', width=0.2, aes(color='black'))+
   scale_x_continuous(breaks=c(16,32,64,128,256,384))+

### 2015-09-22 full/plots/retwis-followers.r

@@ -18,9 +18,7 @@ d <- tryCatch({
 
 r <- d[1,]
 
-save(
-  # ggplot(subset(nbid.hist, x != 0), aes(x=x, weight=y))+
-  ggplot(
+g.followers <- ggplot(
     subset(to.hist(r$server_followers_hist), x > 0),
     aes(x = x, y = y)
   )+
@@ -29,11 +27,11 @@ save(
     scale_x_log10(labels=si.labels())+ #, breaks=c(1,10,100,1000,10000))+
     scale_y_log10(labels=si.labels())+ #, breaks=c(1,10,100,1000,10000,100000))+
     my_theme()
-, 'retwis-followers', w=3.4, h=3.0)
 
-save(
-  # ggplot(subset(nbid.hist, x != 0), aes(x=x, weight=y))+
-  ggplot(
+save(g.followers, 'retwis-followers', w=3.4, h=3.0)
+
+
+g.followers.cdf <- ggplot(
     subset(to.hist(r$server_followers_hist), x > 0),
     aes(x = x, weight = y)
   )+
@@ -42,11 +40,10 @@ save(
     scale_x_log10(labels=si.labels(), breaks=c(10,50,100,1000))+
     scale_y_log10(breaks=c(0.10, 0.5, 1))+
     my_theme()
-, 'retwis-followers-cdf', w=3.4, h=3.0)
+    
+save(g.followers.cdf, 'retwis-followers-cdf', w=3.4, h=3.0)
 
-save(
-  # ggplot(subset(nbid.hist, x != 0), aes(x=x, weight=y))+
-  ggplot(
+g.reposts <- ggplot(
     subset(to.hist(r$server_reposts_hist), x > 0),
     aes(x = x, y = y)
   )+
@@ -55,4 +52,9 @@ save(
     scale_x_log10(labels=si.labels(), breaks=c(1,10,100,1000,10000))+
     scale_y_log10(labels=si.labels(), breaks=c(1,10,100,1000,10000))+
     my_theme()
-, 'retwis-reposts', w=3.4, h=3.0)
+
+save(g.reposts, 'retwis-reposts', w=3.4, h=3.0)
+
+# pdf('retwis-dists.pdf')
+# multiplot(g.followers, g.reposts, cols=2)
+# dev.off()

### 2015-09-22 full/plots/retwis-tput-vs-lat.r

@@ -60,22 +60,22 @@ plot <- function(suffix, d, layers) {
     my_theme()+layers #+theme(legend.position='bottom')
   , 'retwis-tput-vs-lat'+suffix, w=5, h=3)
 
-  # save(
-  #   ggplot(subset(d), aes(
-  #     x = x,
-  #     y = throughput,
-  #     group = x(cc,phasing),
-  #     fill = cc_ph, color = cc_ph, linetype = cc_ph
-  #   ))+
-  #   # geom_point()+
-  #   stat_summary(geom='line', fun.y=mean)+
-  #   expand_limits(y=0)+
-  #   scale_y_continuous(labels=function(x){ x/1000+'k' })+
-  #   facet_wrap(~facet)+ #, scales="free")+
-  #   # cc_scales()+phasing.linetype()+
-  #   cc_ph_scales()+
-  #   my_theme() #+theme(legend.position='bottom')
-  # , 'retwis-explore'+suffix, w=8, h=6)
+  save(
+    ggplot(subset(d, x <= 256), aes(
+      x = x,
+      y = throughput,
+      group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph
+    ))+
+    xlab('Clients')+ylab('Throughput (txn/s)')+
+    stat_summary(geom='line', fun.y=mean)+
+    stat_summary(geom='point', fun.y=mean)+  
+    scale_x_continuous(trans=log2_trans(), breaks=c(8,16,32,64,128,256,384))+
+    scale_y_continuous(labels = function(x){ x/1000+'k' })+
+    expand_limits(x=0, y=0)+
+    facet_wrap(~facet, scales="free")+
+    cc_ph_scales()+
+    my_theme()
+  , 'retwis-tput', w=5, h=3)
 
 }
 

### 2015-09-22 full/plots/rubis-tput-vs-lat.r

@@ -49,18 +49,19 @@ save(
   my_theme()+coord_cartesian(y=c(0,10))
 , 'rubis-tput-vs-lat', w=5, h=3)
 
-# save(
-#   ggplot(subset(d), aes(
-#     x = x,
-#     y = throughput,
-#     group = x(cc,phasing),
-#     fill = cc_ph, color = cc_ph, linetype = cc_ph
-#   ))+
-#   # geom_point()+
-#   stat_summary(geom='line', fun.y=mean)+
-#   expand_limits(y=0)+
-#   scale_y_continuous(labels=function(x){ x/1000+'k' })+
-#   facet_wrap(~facet)+ #, scales="free")+
-#   cc_ph_scales()+
-#   my_theme()
-# , 'rubis-explore', w=8, h=6)
+save(
+  ggplot(subset(d, x <= 256), aes(
+    x = x,
+    y = throughput,
+    group = cc_ph, fill = cc_ph, color = cc_ph, linetype = cc_ph
+  ))+
+  xlab('Clients')+ylab('Throughput (txn/s)')+
+  stat_summary(geom='line', fun.y=mean)+
+  stat_summary(geom='point', fun.y=mean)+  
+  scale_x_continuous(trans=log2_trans(), breaks=c(8,16,32,64,128,256,384))+
+  scale_y_continuous(breaks = c(0, 5000, 10000), labels = si.labels())+
+  expand_limits(x=0, y=0)+
+  facet_wrap(~facet, scales="free_x")+
+  cc_ph_scales()+
+  my_theme()
+, 'rubis-tput', w=5, h=3)

### 2015-09-23 papoc/paper.Rmd
fix acm number
fix broken ref to bloom -> bloomfilter

@@
   year: 2015
-  copyrightdata: 978-1-4503-3238-5/15/04
+  copyrightdata: 978-1-4503-3537-9/15/04
 
@@
 
-**Probabilistic data types** such as *bloom filters* [@bloom], *hyperloglogs* [@hyperloglog], and *count-min sketches* [@countminsketch] trade off accuracy (within fixed bounds) for better performance or storage.
+**Probabilistic data types** such as *bloom filters* [@bloomfilter], *hyperloglogs* [@hyperloglog], and *count-min sketches* [@countminsketch] trade off accuracy (within fixed bounds) for better performance or storage.
 Twitter's streaming analytics system [@summingbird] and many machine learning algorithms leverage these to handle high data volume, and we expect similar benefit.
